---
title: "Classifying County-Level Wildfire Smoke"
author: "Ryan Gan"
date: "2018-03-02"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

I need a way to efficiently classify if a county was impacted by smoke. I plan to use population-weighted PM~2.5~, air quality index, temperature, and HMS for each county to do this. 

I also want to assemble data that I may consider a "true" indications of smoke, where I'll look at news reports and public health warnings that specifically mention smoke has caused harmful air quality days. I would then like to ocmpare how accurate my classifiers are against these measures of truth.

```{r setup}
# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, 
                      warning=FALSE, message=FALSE)

# libary
library(tidyverse)
```

Import 2010-2015 county PM and air quality data.

```{r pm_import}
pm <- read_csv("../../data/smoke/1015-county_popwt_pm.csv") %>% 
  mutate(hms_percent = hms*100,
         smoke10_hms = ifelse(smoke10 == 1 & hms > 0.50, 1, 0),
         smoke15_hms = ifelse(smoke15 == 1 & hms > 0.50, 1, 0),
         smokewv_hms = ifelse(smoke_wave == 1 & hms > 0.50, 1, 0),
         aqi_hms = ifelse(aqi_cat != "Good" & hms > 0.50, 1, 0))

xtabs(~smoke10_hms + smoke10, data = pm)

xtabs(~smokewv_hms + smoke15_hms, data = pm)
xtabs(~aqi_cat, data = pm)
xtabs(~aqi_hms + smoke15_hms, data = pm)
```

```{r aqi}
ggplot(pm, aes(x=date,y=aqi)) +
  geom_point() +
  facet_wrap(~state) +
  ylim(0, 1000)
```

```{r}
ggplot(pm, aes(x=hms, y=aqi)) +
  geom_point() +
  facet_wrap(~month) +
  ylim(0,500)

ggplot(pm, aes(x=hms, y=temp_f)) +
  geom_point() +
  facet_wrap(~month) 

ggplot(pm, aes(x=hms, y=pm_krig)) +
  geom_point() +
  facet_wrap(~month) 

mod <- lm(pm_krig ~ hms_percent , data=pm)
summary(mod)

xtabs(~smoke15 + aqi_cat, data=pm)
```

```{r colorado_4mile}
check <- pm %>% 
  filter(state == "Colorado" & year == 2010 & fips == "08013") %>% 
  select(date, pm_krig, pm_smk, aqi, hms_percent) %>% 
  gather("exposure", "value", -date) %>% 
  filter(date >= "2010-09-01" & date <= "2010-10-01")

ggplot(check, aes(x=date, y=value)) +
  geom_point() +
  facet_wrap(~exposure, scales = "free_y") +
  theme_minimal()

flag_4mile <- pm %>% 
  filter(state == "Colorado") %>% 
  filter(date >= "2010-09-01" & date <= "2010-10-01") %>% 
  mutate(flag = ifelse(hms > 0 & pm_smk > 5, 1, 0)) %>% 
  filter(flag ==1)
```

Flagging counties that were likely affected by big fires listed in the grant. To be flagged, the county must be in a specific state listed, within the date range listed, and have an AQI of unhealthy for sensitive groups or higher.

```{r flag_fire_smoke}
# 4 mile
flag_4mile <- pm %>% 
  filter(state == "Colorado") %>% 
  filter(date >= "2010-09-06" & date <= "2010-09-17") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "four_mile") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)

# rim fire
flag_rim <- pm %>% 
  filter(state %in% c("California", "Nevada")) %>% 
  filter(date >= "2013-08-17" & date <= "2013-10-25") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "rim") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)

# central idaho
flag_cen_id <- pm %>% 
  filter(state %in% c("Idaho")) %>% 
  filter(date >= "2013-08-12" & date <= "2013-08-21") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "cen_idaho") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)

# douglas complex
flag_doug <- pm %>% 
  filter(state %in% c("Oregon")) %>% 
  filter(date >= "2013-07-25" & date <= "2013-09-05") %>% 
  mutate(flag = ifelse(hms > 0.50 &
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "douglas_complex") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)

# 2012 washington fires
flag_wen <- pm %>% 
  filter(state %in% c("Washington", "Idaho", "Oregon")) %>% 
  filter(date >= "2012-09-07" & date <= "2012-09-30") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "wenatchee") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)

#  2012 idaho
flag_id12 <- pm %>% 
  filter(state %in% c("Montana", "Idaho", "Wyoming")) %>% 
  filter(date >= "2012-08-11" & date <= "2012-09-16") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "idaho2012") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)

# high park
flag_hp <- pm %>% 
  filter(state %in% c("Colorado", "Wyoming")) %>% 
  filter(date >= "2012-06-09" & date <= "2012-06-30") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "high_park") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)

# fire flags
smoke_train <- mget(grep("flag", names(.GlobalEnv), value=TRUE)) %>% 
  map_dfr(., bind_rows) 
```

Now that I have possible smoke-impacted dates for county, I'd like to find referent periods. 
Take counties that likely were impacted by smoke and bind it back in to the time series for those counties.

```{r}
# find unique dates
smoke_dates <- smoke_train %>% 
  group_by(fips, date) %>% 
  summarise(n = n()) %>% 
  select(fips,date) %>% 
  as.matrix()

# find ref dates to output
fips_ref_dates <- apply(smoke_dates, 1, function(x){
  before <- as.Date(x[2], format = "%Y-%m-%d") %m-% months(1:12)
  after <- as.Date(x[2], format = "%Y-%m-%d") %m+% months(1:12)
  ref_date <- as.character(c(after, before)) 
  fips <- rep(x[1], length(ref_date))
  fips_ref_date <- data.frame(fips, ref_date)
  return(fips_ref_date)
      }
    ) %>% 
  rbind_list() %>% 
  mutate(date = as.Date(ref_date))%>% 
  select(fips, date) %>% 
  left_join(pm, by = c("fips", "date")) %>% 
  mutate(flag = ifelse(hms < 0.10 & smoke15 == 0, 0, 1))

glimpse(fips_ref_dates)

fips_ts <- smoke_train %>% 
  select(fips, date, flag, fire) %>% 
  left_join(pm, by = c("fips", "date")) %>% 
  bind_rows(fips_ref_dates) %>% 
  arrange(fips, date) %>% 
  mutate(smoke_true = if_else(flag == 1, 1, 0),
         month = as.factor(month))

glimpse(fips_ts)
```

```{r test_mod}
train <- fips_ts %>% 
  filter(state != "California")

test <- fips_ts %>% 
  filter(state == "California")



library(randomForest)

summary(train)
xtabs(~smoke_true, data = train)

# model without aqi
rf_mod <- randomForest(as.factor(smoke_true) ~ pm_krig + bg_pm + 
  pm_smk + hms + temp_f + month, 
  data=train, importance=TRUE, ntree=500)

varImpPlot(rf_mod)

# predict cali fires
cali_pred <- predict(rf_mod, test)

# join with test
test2 <- cbind(test, cali_pred)

xtabs(~cali_pred + smoke_true, test2)
xtabs(~smoke_wave + smoke_true, test2)
xtabs(~smoke15 + smoke_true, test2)

colorado_2015 <- pm %>% 
  filter(state == "Colorado" & year == 2015) %>% 
  mutate(month = as.factor(month))

glimpse(colorado_2015)
colorado_2015$smoke_pred <- predict(rf_mod, colorado_2015)

xtabs(~ smoke_pred + smoke_wave, colorado_2015)

co_smk_check <- colorado_2015 %>% 
  filter(smoke_pred == 1) %>% 
  filter(fips == "08069")

xtabs(~aqi_cat, co_smk_check)
```

