---
title: "Classifying County-Level Wildfire Smoke"
author: "Ryan Gan"
date: "2018-03-02"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Introduction
*Problem: I need a binary classifier that indicates that air quality in a county was affected by smoke.*

I would like to find a way to create a binary classifier of smoke with relatively reasonable accuracy with decent sensitivity and specificity. Up to this point, I've used straight forward methods like subtracting estimated PM~2.5~ from median or mean estimated background estimates, and then further defining smoke as this difference greater than 0, 10, and 15 ug/m^3^ respectively. I've even applied Coco Liu's smoke wave day where I've applied the 98^th^ percentile of the PM~2.5~ distribution for counties in the Western US. One key difference with my approach is that I am using smoothed (via krig) PM~2.5~ values from surface site monitors, where Coco used chemical transport model estimates of PM~2.5~ specific to smoke. 

## Purpose/Aim
It's unclear how accurate any of these approaches to classify smoke at some spatial unit (county in this case) really are. The goal of this document is to draft a outline or approach to classify binary smoke exposure and to assess the sensitivity, specificity, and overall accuracy of each method.

### Methods Overview
I need a way to efficiently classify if a county was impacted by smoke. I plan to use the time series of population-weighted PM~2.5~, air quality index, temperature, and HMS for each county I created to do this. 

The first step is to assemble data that I consider a "true" indications of smoke. This is nuanced, where a county can experience smoke that impacts one portion of the county but not the other portions. It also might impact the portion where most people live. The aggregated PM~2.5~ and smoke values I have clearly don't have this level of detail. However, I think there is still value to understand the differences in my binary classifier. The first step is to determine counties and times that were very likely to be impacted by smoke, and areas unlikely to be affected by smoke, while still maintaining some variability in the predictors so that we can create a decent estimator of smoke. This will serve as our "truth" examples.

```{r setup}
# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, 
                      warning=FALSE, message=FALSE)

# libary
library(tidyverse) # tidyverse
```

Import 2010-2015 county PM and air quality data. I have coded in some binary smoke variables already. I first calculated differences between estimated daily PM~2.5~ for a county and median background PM~2.5~ for the season when smoke wasn't present in the atmospheric column. I've coded differences as >0, >5, >10, and >15 as smoke0, smoke5, smoke10, and smoke15 respectively. I've also coded the smoke wave variable, which is at least 2 days of PM~2.5~ levels > the 98^th^ percentile (22.3 ug/m^3).

```{r pm_import}
# import
pm <- read_csv("../../data/smoke/1015-county_popwt_pm.csv") %>% 
  mutate(hms_percent = hms*100,
         smoke10_hms = ifelse(smoke10 == 1 & hms > 0.50, 1, 0),
         smoke15_hms = ifelse(smoke15 == 1 & hms > 0.50, 1, 0),
         smokewv_hms = ifelse(smoke_wave == 1 & hms > 0.50, 1, 0),
         aqi_hms = ifelse(aqi_cat != "Good" & hms > 0.50, 1, 0)) %>% 
  rename(pm_diff = pm_smk)

# calculating state medians for plots
median_vals <- pm %>% 
  group_by(state, date) %>% 
  summarise_at(vars(pm_krig:aqi, hms_percent), median, na.rm=T)
```

### PM Plot
Plotting kriged PM~2.5~ estimates of each fips for each state. I've also plotted state median value as a red line.

```{r pm_plot}
plot <- ggplot(pm, aes(x=date,y=pm_krig)) +
  geom_point() +
  geom_line(data=median_vals, aes(x=date, y=pm_krig), color = "red") +
  facet_wrap(~state) +
  theme_minimal()

print(plot)
```

California has periodic spikes in PM~2.5~. This could be wildfire smoke, but could there do appear to be peaks in winter and other times when wildfire smoke is less common, suggesting other sources. The peak in Nevada around summer of 2013 is likely the Rim fire. Oregon also has some peaks that are likely fires. For example, the Douglas-Complex fire of 2013 looks like it's in the right place. Washington also has peaks that I know are associated with the 2012 Wenatchee fire and 2015 fires. Colorado and Wyoming have less-pronounced peaks. Could this possibly be due to lower performance of krig in places with sparse monitors?

### AQI Plot
Plotting air quality index (AQI) for each fips in each state. Plotting median AQI value as red line as well. I download AQI data for each county from the EPA AirNow website. There was an AQI value of over 10,000. I'm not sure this is real, so I've clipped the y axis limit to 500. There were missing values for counties and days, which I'm guessing is common. Red line is the state median time series.

```{r aqi}
plot <- ggplot(pm, aes(x=date,y=aqi)) +
  geom_point() +
  geom_line(data=median_vals, aes(x=date, y=aqi), color = "red") +
  facet_wrap(~state) +
  ylim(0, 500) +
  theme_minimal()

print(plot)
```

### Temperature (Farenheit) Plot
Plotting temperature in Farenheit for each county in each state. Temperature data was collected from the NOAA reanalysis data repository. Red line is the state median time series.

```{r temp_plot}
plot <- ggplot(pm, aes(x=date,y=temp_f)) +
  geom_point() +
  geom_line(data=median_vals, aes(x=date, y=temp_f), color = "red") +
  facet_wrap(~state) +
  theme_minimal()

print(plot)
```

### HMS Plot
Plotting percent of county with smoke in the atmospheric colum assessed using HMS. Red line is the median precent for the state on that day.

```{r hms_plot}
plot <- ggplot(pm, aes(x=date,y=hms_percent)) +
  geom_point() +
  geom_line(data=median_vals, aes(x=date, y=hms_percent), color = "red") +
  facet_wrap(~state) +
  theme_minimal()

print(plot)
```

Hmm there are some days where a lot of counties were impacted by smoke I guess. I'm not sure if the median value is that useful here.

### Test Case: Four-Mile Canyon Fire
On September 7^th^, 2010, there was an air quality warning specifically related to smoke from four-mile canyon fire that was publicized in the news. I want to look at the time series for these variables for Boulder county (fips == 08013) during September of 2010.

```{r colorado_4mile}
# subset
check <- pm %>% 
  filter(state == "Colorado" & year == 2010 & fips == "08013") %>% 
  select(date, pm_krig, pm_diff, aqi, hms_percent) %>% 
  gather("exposure", "value", -date) %>% 
  filter(date >= "2010-09-01" & date <= "2010-10-01")
# plot
plot <- ggplot(check, aes(x=date, y=value)) +
  geom_point() +
  facet_wrap(~exposure, scales = "free_y") +
  xlab("2010") +
  theme_minimal()

print(plot)
```

AQI, PM krig, PM difference, are all elevated on Septebmer 7th. HMS precent is a little more confusing I think. The county has smoke present on September 7th, but there are other days with smoke in the column where air quality and PM levels are not as extreme, suggesting smoke overhead, but that it did not impact air quality in the area.

However, this leads me to my first-draft case definition of a "true" day impacted by smoke. 

### "True" Smoke Days
I'm going to use the fires listed in the NASA grant. I'll use the states and time periods listed in the grant to filter the larger PM data to these days and areas. I will then flag counties that were likely affected by big fires listed in the grant. To be flagged, the county must be in a specific state listed, within the date range listed, and have an AQI of unhealthy for sensitive groups or higher. Note that this will exclude counties and days without an AQI measurement, but I think this is a decent classifier of "true". I've also required that the county be covered by at least 50% of smoke. I can see shortcomings to this also. I'm open to feedback to make this better since this is not my expertise.

Flagging 4 mile canyon fire in Colorado, 2010-09-06 to 2010-09-17.

```{r flag_4mile}
# 4 mile
flag_4mile <- pm %>% 
  filter(state == "Colorado") %>% 
  filter(date >= "2010-09-06" & date <= "2010-09-17") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "four_mile") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)
```

Flagging Rim fire in Nevada and California, from 2013-08-17 to 2013-10-25.

```{r flag_rim}
# rim fire
flag_rim <- pm %>% 
  filter(state %in% c("California", "Nevada")) %>% 
  filter(date >= "2013-08-17" & date <= "2013-10-25") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "rim") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)
```

Flagging central Idaho fires, 2013-08-12 to 2013-08-21.
```{r flag_idaho}
# central idaho
flag_cen_id <- pm %>% 
  filter(state %in% c("Idaho")) %>% 
  filter(date >= "2013-08-12" & date <= "2013-08-21") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "cen_idaho") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)
```

Flagging Douglas-Complex fires in Oregon, 2013-07-25 to 2013-09-25.
```{r flag_doug}
# douglas complex
flag_doug <- pm %>% 
  filter(state %in% c("Oregon")) %>% 
  filter(date >= "2013-07-25" & date <= "2013-09-05") %>% 
  mutate(flag = ifelse(hms > 0.50 &
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "douglas_complex") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)
```

Flagging Wenatchee-Complex fires in Washington, Idaho, and Oregon, 2012-09-07 to 2012-09-30.
```{r wen_fires}
# 2012 washington fires
flag_wen <- pm %>% 
  filter(state %in% c("Washington", "Idaho", "Oregon")) %>% 
  filter(date >= "2012-09-07" & date <= "2012-09-30") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "wenatchee") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)
```

Flagging Idaho fires in 2012-08-11 to 2012-09-16.
```{r flag_idaho12}
#  2012 idaho
flag_id12 <- pm %>% 
  filter(state %in% c("Montana", "Idaho", "Wyoming")) %>% 
  filter(date >= "2012-08-11" & date <= "2012-09-16") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "idaho2012") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)
```

Flagging High Park fire in Colorado and Wyoming, 2012-06-09 to 2012-06-30.
```{r highpark}
# high park
flag_hp <- pm %>% 
  filter(state %in% c("Colorado", "Wyoming")) %>% 
  filter(date >= "2012-06-09" & date <= "2012-06-30") %>% 
  mutate(flag = ifelse(hms > 0.50 & 
    !(aqi_cat %in% c("Good", "Moderate", NA)), 1, 0)) %>% 
  filter(flag ==1) %>% 
  mutate(fire = "high_park") %>% 
  select(state, fips, date, hms, pm_krig, aqi_cat, flag, fire)
```

Binding all the flagged fires together to create one dataframe.
```{r flag_fires}
# fire flags
smoke_train <- mget(grep("flag", names(.GlobalEnv), value=TRUE)) %>% 
  map_dfr(., bind_rows) 
```

### "True" No Smoke
Now that I have smoke-impacted dates for county, I'd like to find referent periods where smoke was unlikely. I'm a little less clear on how to do this, but here it goes. I'm going to take counties that were likely impacted by smoke and bind it back in to the time series for those counties giving my a a time series for those flagged counties for the entire time period.

```{r bind_smoke}
# find unique dates
smoke_dates <- smoke_train %>% 
  group_by(fips, date) %>% 
  summarise(n = n()) %>% 
  select(fips,date) %>% 
  as.matrix()
```

I'm then going to take observations from the same time series for a year before and a year after the flagged smoke day. I'm then going to consider a smoke day where HMS over the county is less than 10 percent. This could definately be adjusted.

Note I tried county less than 0 percent and this was by far the single biggest predictor of smoke in algorithms I used later and actually didn't seem to perform well with scenarios where smoke was overhead, but ground-level air quality seemed fine (it would predict smoke).
```{r ref_dates}
# find ref dates to output
fips_ref_dates <- apply(smoke_dates, 1, function(x){
  before <- as.Date(x[2], format = "%Y-%m-%d") %m-% months(1:12)
  after <- as.Date(x[2], format = "%Y-%m-%d") %m+% months(1:12)
  ref_date <- as.character(c(after, before)) 
  fips <- rep(x[1], length(ref_date))
  fips_ref_date <- data.frame(fips, ref_date)
  return(fips_ref_date)
      }
    ) %>% 
  rbind_list() %>% 
  mutate(date = as.Date(ref_date))%>% 
  select(fips, date) %>% 
  left_join(pm, by = c("fips", "date")) %>% 
  mutate(flag = ifelse(hms < 0.25 & pm_diff < 10, 0, 1))
```

Now I'm going to extract flagged smoke and non-smoke dates and build a fips time series data to train and validate prediction algorithms on.

```{r smoke}
fips_ts <- smoke_train %>% 
  select(fips, date, flag, fire) %>% 
  left_join(pm, by = c("fips", "date")) %>% 
  bind_rows(fips_ref_dates) %>% 
  arrange(fips, date) %>% 
  mutate(smoke_true = if_else(flag == 1, 1, 0),
         month = as.factor(month))
```

### Sensitivity and Specificity of Existing Binary Smoke Classifiers
Before I apply an algorithm, I'm going to see how well our existing classifiers perform against the "true" smoke measure. 

```{r accuracy}
acc_results <- c("smoke0", "smoke5", "smoke10", "smoke15", "smoke_wave",
                 "smoke10_hms", "smoke15_hms", "smokewv_hms", "aqi_hms") %>% 
  map_dfr(., function(x){
    outcome <- x
    tab <- xtabs(as.formula(paste0("~",x, "+ smoke_true")), fips_ts)
    sn <- as.numeric(round(prop.table(tab, 2)[4],3)*100)
    sp <- as.numeric(round(prop.table(tab, 2)[1],3)*100)
    acc <- as.numeric(round((sn+sp)/2,1))
    ppv <- as.numeric(round(prop.table(tab, 1)[4],3)*100)
    npv <- as.numeric(round(prop.table(tab, 1)[1],3)*100)
    measures <- data.frame(cbind(outcome, sn, sp, acc, ppv, npv)) %>% 
      mutate_at(vars(sn:npv), funs(as.numeric(as.character(.))))
    return(measures)
})

acc_results
```


I'm going to split the data in to a trainning and validation set using the caret package.


```{r test_mod}
train <- fips_ts %>% 
  

test <- fips_ts %>% 
  filter(state == "California")


library(randomForest) # randomforest package


summary(train)
xtabs(~smoke_true, data = train)

# model without aqi
rf_mod <- randomForest(as.factor(smoke_true) ~ pm_krig + bg_pm + 
  pm_smk + hms + temp_f + month, 
  data=train, importance=TRUE, ntree=500)

varImpPlot(rf_mod)



# predict cali fires
cali_pred <- predict(rf_mod, test)

# join with test
test2 <- cbind(test, cali_pred)

xtabs(~cali_pred + smoke_true, test2)
xtabs(~smoke_wave + smoke_true, test2)
xtabs(~smoke15 + smoke_true, test2)

colorado_2015 <- pm %>% 
  filter(state == "Colorado" & year == 2015) %>% 
  mutate(month = as.factor(month))

glimpse(colorado_2015)
colorado_2015$smoke_pred <- predict(rf_mod, colorado_2015)

xtabs(~ smoke_pred + smoke_wave, colorado_2015)

co_smk_check <- colorado_2015 %>% 
  filter(smoke_pred == 1) %>% 
  filter(fips == "08069")

xtabs(~aqi_cat, co_smk_check)
```

