---
title: "Intersection of Grid and County Shapefiles"
author: "Ryan Gan"
date: "2018-01-17"
output: html_document
---

## Purpose

This markdown script calculates the proportion of intersection between the 2015 grid shape file and county shape file. This grid is different from the 2010-2015 Western US grid for kriged estimates.

## Setup

Loading tidyverse for general data wrangling and plotting and sf package for spatial calculations.

```{r setup, include=F}
# libraries used
library(tidyverse)
library(sf)

# general knitr options
knitr::opts_chunk$set(fig.width = 7, fig.height = 5, 
                      message = F, warning = F)
```

## Shapefiles

Reading in the grid and county shape files as simple features. 

Also consider county boundaries. I'm sure they don't change much, but may not hurt to use 2015 shape files.

```{r read_sf}
# grid path
grid_path <- paste0("../data/shapefile/2015-wus_grid")
# read grid sf
grid_sf <- st_read(dsn = grid_path, layer = "wus_grid2015-ko") 

# county path
county_path <- paste0("../data/shapefile/us_county")
# read county sf
county_sf <- st_read(dsn = county_path, layer = "us_county")

# assign crs of county_sf to grid_sf
wgs84 <- st_crs(county_sf)
# assign crs
st_crs(grid_sf) <- wgs84
```

### Plotting overlay of grid and county shapefiles

Subset US county shape file to only Western US states and counties and plot it to make sure it looks right.

```{r western_subset}
# STATE FIPS I want to subset
western_state_fp <- c("04","06","08","16","53","41","32","49","56",
                      "35", "30")

# subset simple features
western_county_sf <- county_sf %>% 
  filter(STATEFP %in% western_state_fp) %>% 
  mutate(FIPS = paste0(STATEFP, COUNTYFP))

# plot
plot <- ggplot(western_county_sf) +
  geom_sf(color = "red") +
  geom_sf(data=grid_sf, aes(), alpha = 0.0) +
  theme_minimal()

plot
```

## Grid ID

Checking how WRFGRID ID is labeled. It looks like this starts in the bottom left hand corner, goes bottom to top, then left to right. This is not necissarily how values are extracted from the NetCDF file. I will need to be aware of this.

```{r wrfgrid_label}
# subset to first 10 WRFGRID IDs
grid_subset <- grid_sf %>% 
  slice(c(1:10, 152:162))
# name of grid
# make a dataframe with rounded proportion values
grid_names <- grid_subset %>% 
  group_by(WRFGRID_ID) %>% 
  mutate(lon = unlist(st_centroid(geometry))[1],
         lat = unlist(st_centroid(geometry))[2]) %>% 
  ungroup() %>% 
  select(WRFGRID_ID, lon, lat)
# convert simple featrues to dataframe
st_geometry(grid_names) <- NULL

# plot
plot <- ggplot(grid_subset) +
  geom_sf() +
  geom_text(data=grid_names, aes(x=lon, y=lat,label= WRFGRID_ID)) +
  theme_minimal()

plot
```

Projection looks good. I want to calculate proportion intersection for each county in these states. I will need to make sure I extract values from the NetCDF file in this way otherwise my population-weight calculations will be off.

## Calculation of proportion intersect

Going to calculate in a for loop because I haven't figured out a good apply or purrr way to do this. The loop only takes about 2.6 minutes, which is not that much time.

```{r propotion_intersect}
# create empty dataframe
prop_int_tibble <- grid_sf$WRFGRID_ID %>% 
  tibble() %>% 
  rename(grid_id = ".")

# start compute time
start_time <- Sys.time()

for(i in 1:length(western_county_sf$FIPS)){
  # subset county to find intersect
  county <- slice(western_county_sf, i)
  # extract fips number for variable name
  fips_id <- paste0("fips_", county$FIPS)
  # subset grid cells that touch any part of the county
  grid <- grid_sf[county,]
  # subset the intersected polygon
  inter_poly <- st_intersection(grid, county) %>% 
    # filter only to polygon or multipolygon type 
    # to avoid errors with point or line types
    filter(st_is(., c("POLYGON", "MULTIPOLYGON"))==T)
  # filter grid ids to only grids in the inter_poly object
  grid2 <- grid %>% filter(WRFGRID_ID %in% inter_poly$WRFGRID_ID) 
  # find proportion intersect with original grid
  prop_int <- as.numeric(st_area(inter_poly)/st_area(grid2))
  # subset grid id
  grid_id <- grid2$WRFGRID_ID
  # make a tibble
  county_grid_int <- tibble(grid_id, prop_int) %>% 
    set_names(c("grid_id", fips_id))
  # join with full tibble
  prop_int_tibble <- prop_int_tibble %>% 
    left_join(county_grid_int, by = "grid_id")
} # end loop

# compute time
stop_time <- Sys.time()
compute_time <- stop_time - start_time
# run time
compute_time
```

### Check of intersection in LA county

Checking grid intersection calculations for a specific county to see if calculations look reasonable. I'll use Los Angles county since I've made these calculations before for the Bluesky grid.

```{r la_county_check}
# subset la county (FIPS 06037)
la_county <- western_county_sf %>% 
  filter(FIPS == "06037")

# subset proportion intersect values for LA county
la_prop_int <- prop_int_tibble %>% 
  select(grid_id, fips_06037) %>% 
  filter(!is.na(fips_06037))

# subset la grid
la_grid <- grid_sf %>% 
  filter(WRFGRID_ID %in% la_prop_int$grid_id) %>% 
  left_join(la_prop_int, by = c("WRFGRID_ID" = "grid_id")) %>% 
  rename(prop_int = fips_06037)

# make a dataframe with rounded proportion values
grid_names <- la_grid %>% 
  mutate(proportion = round(prop_int,2)) %>% 
  group_by(WRFGRID_ID) %>% 
  mutate(lon = unlist(st_centroid(geometry))[1],
         lat = unlist(st_centroid(geometry))[2]) %>% 
  ungroup() %>% 
  select(WRFGRID_ID, lon, lat, proportion)
# convert simple featrues to dataframe
st_geometry(grid_names) <- NULL

# plot grid overlayed on county
plot <- ggplot(la_county) + 
  geom_sf() +
  geom_sf(data=la_grid, aes(fill=prop_int), alpha=0.7) +
  geom_text(data=grid_names, aes(x=lon, y=lat, label=proportion), 
            size = 2.5, angle=45) +
  theme_bw()

plot
```

Plot looks good of intersections look good. I'm going to do some final manipulations on the proportion intersect data frame and save a permanent file.

### Writing proportion intersect file

File will be used for calculations of county population-weighted PM.

```{r save_file, eval=F}
# replace NAs with 0s
wus_prop_int <- prop_int_tibble %>% 
  rename(WRFGRID_ID = grid_id) %>% 
  mutate_all(funs(replace(., is.na(.), 0)))

# save file
write_csv(wus_prop_int, paste0("./data/smoke/2015-smoke_propint_wus.csv"))
```