---
title: "Regrid of NOAA reanalysis temp data to krig PM grid"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# Introduction

This R notebook contains code to regrid the NOAA reanalysis air temperature at 2 meters to match the grid used for the kriged estimates of PM~2.5~.

```{r setup}
# libraries used
library(tidyverse) 
library(ncdf4) # package for working with .nc files
library(raster) # raster package
library(rgdal) # spatial transformation

# general knitr options
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

## PM~2.5~ NetCDF
Reading in krig PM~2.5~ netcdf file.

```{r pm_nc_read}
# read pm nc
pm_nc_path <- paste0("../../data/smoke/krigedPM25_06-15/krigedPM25_2015.nc")
pm_nc <- nc_open(pm_nc_path)
# summary of krig nc
pm_nc
```

Extract latitude and longitude coordinates for PM grid.

```{r pm_latlon}
# extract lat and lon
pm_lat <- ncvar_get(nc = pm_nc, varid = "lat")
pm_lon <- ncvar_get(nc = pm_nc, varid = "lon")
pm_val <- ncvar_get(nc = pm_nc, varid = "PM25")[,,1]

# combine in dataframe
pm_location_val <- as_data_frame(cbind(as.vector(pm_lon), as.vector(pm_lat),
                             as.vector(pm_val))) %>% 
  rename(lon = V1, lat = V2, pm = V3) %>% 
  mutate(grid = as.factor("pm"))
```

### Plot of PM~2.5~ Grid
Plotting PM~2.5~ values on 2015-01-01. Looks like there was some bad air pollution in Souther California. Projection may be a bit off, with this quick map. I'm using map outline from map package so I can't change the projection without a lot of work, but I will make sure I change the projection so that it looks correct on any final product.

```{r pm_plot}
# extract usa polygon from map package
usa_map <- maps::map("usa")
# plot pm values over us shape
ggplot(pm_location_val, aes(x=lon, y=lat)) +
  geom_point(aes(color=pm)) +
  geom_polygon(data = usa_map, aes(x=long,y=lat, group=group), fill=NA,
               color="red") +
  coord_fixed(1.3) +
  theme_minimal()
```

## Temperature NetCDF
Reading temperature data from reanalysis netcdf file.

```{r temp}
# temp nc_path
temp_nc_path <- paste0("../../data/smoke/air.2m.2015.nc")
temp_nc <- nc_open(temp_nc_path)
# temp nc summary
temp_nc
```

Extract temperature grid latitude and longitude coordinates. It looks like this grid is a more extensive range than the PM grid, which is just the continental US.

```{r temp_latlon}
# extract temp lat lon
t_lat <- ncvar_get(nc = temp_nc, var = "lat")
t_lon <- ncvar_get(nc = temp_nc, var = "lon")
# extract air temp on jan 1st
t_temp <- ncvar_get(nc  =temp_nc, var = "air")[,,1]
# combine in dataframe
t_location <- as_data_frame(cbind(as.vector(t_lon), as.vector(t_lat),
                                  as.vector(t_temp))) %>% 
  rename(lon = V1, lat = V2, temp_k = V3) %>% 
  mutate(grid = as.factor("temp"))
```

### Plot of Temperature Grid
Plotting gridded Temperature values (in Kelvin) on 2015-01-01. The domain is so large that I can plot as points with decent enough resolution. It looks like you can see the outline of North America as the air temperature (at 2 meters) is much cooler than the air just above the oceans at lower latitudes. Cool. Also, it looks like some points are missing temp values. The longitude range also extends in to the Eastern Hemisphere. This matches what NOAA shows.

https://www.esrl.noaa.gov/psd/data/narr/format.html
```{r kelvin_plot}
ggplot(t_location, aes(x=lon,y=lat)) +
  geom_point(aes(color=temp_k), alpha= 0.8) +
  geom_polygon(data = usa_map, aes(x=long,y=lat, group=group), 
               fill=NA, color="red")  +
    coord_fixed(1.3) +
  theme_minimal()
```

### Plot of Grid Extents
I want to know the extent of the two grids.

Plotting the centroids of grids for the PM~2.5~ grid (red color) and the temperature grid (blue color). The temperature data extends almost to the artic circle and contains part of the eastern hemisphere. The PM grid box is only over the continental US.

```{r grid_extent}
ggplot(t_location, aes(x=lon,y=lat)) +
  geom_point(size = 0.1, alpha = 0.2, color = "blue") +
  geom_point(data = pm_location_val, aes(x=lon,y=lat), 
             size = 0.1, color = "red") +
  theme_minimal()
```

### Subset Grid Centroids 

Points are hard to see on the extent map so I'm going to subset to a smaller spatial area: lat 40 to 50 degrees north, and lon -100 to -110 degrees west. Plotting centroid points of the two grids. It looks like the temperature grids are further spaced (lower resolution) compared to PM grid.

```{r point_subset}
# bind two grids together
grid_loc <- rbind(pm_location_val[,c(1:2,4)], t_location[,c(1:2,4)])

# subset to a more managable viewing space
grid_sub <- grid_loc %>% 
  filter(lon <= -100 & lon >= -110 & lat <= 50 & lat >=40)

# plot
ggplot(grid_sub, aes(x=lon,y=lat, color = grid)) +
  geom_point(size =1) +
  scale_color_manual(values = c("red", "blue")) +
  theme_minimal()
```

## Regridding Temperature Values to Match PM Grid

I will need to produce county-level estimates of temperature weighted by population. That means I will need to regrid population or temperature. Bonne is going to regrid population for me, so I will regrid temperature.I think the most straight-forward method would be some sort of smoothing/interpolation method, but I also think this could be computationally expensive. I think it's also possible to do a nearest neighbor thing where I've identify the temperature centroid nearest to a PM centroid and just assign it that way. 

After some investigation, I've found that interpolation is way too computationally expensive. And after further thought, I already have a smoothed surface of temperature. I've decided that the most efficient way is to just extract values of the temperature raster to the coordinate points of the PM grid.

First, I want to see if I can subset the temperature coordinates to just the extent of the PM data.

```{r us_temp}
# setting lat lon bounds
lon_min <- min(pm_location_val$lon)
lon_max <- max(pm_location_val$lon)
lat_min <- min(pm_location_val$lat)
lat_max <- max(pm_location_val$lat)

us_temp <- t_location %>% 
  filter(lon >= lon_min & lon <= lon_max & lat >= lat_min & lat <= lat_max)
```

Plot of 2015-01-01 temperature for the US. Red points are the points from the PM grid I want to interpolate to.
```{r us_temp_plot}
ggplot(data=pm_location_val, aes(x=lon,y=lat)) +
  geom_point(color="red", size=0.1) +
  geom_point(data=us_temp, aes(x=lon,y=lat,color=temp_k), 
             size=3, alpha=0.5) +
  scale_color_gradient(low="white", high="blue") +
  theme_minimal()
```

### Creating Temperature Raster from NetCDF

The raster package can read in netcdf files, however the lambert projection was giving me troubles. It looked like it was plotting in meters rather than degrees, but I couldn't convert or transform the spatial projection and units correctly. Instead, I'm building 

Going to try to use the raster package to extract values from the temperature raster to the spatial point coordinates of the PM grid.

### Extracting Coordinates
I will need both points. I will need the PM coordinates to extract values from the temperature raster to. I will need the temperature coordinates to recreate the raster by latitude/longitude in degrees rather than by Lambert projection in meters.
```{r extract_coords}
# extract pm coords
pm_coords <- matrix(cbind(as.vector(pm_lon), as.vector(pm_lat)), ncol=2)
# extract temp cooreds
temp_coords <- matrix(cbind(as.vector(t_lon), as.vector(t_lat)), ncol=2)
```

### Create Empty Raster
Define the extent of the temperature raster I want to create.
```{r define_extent}
# i need to rebuild temp raster with appropriate lat lon coords
# set up empty extent
e <- extent(temp_coords[,1:2])
e
```
Create an empty raster based on the defined extent, and by the dimensions of the original temperature grid.
```{r create_empty_raster}
# create empty raster of point centroids I want to assign temp values to
r <- raster(e, nrow=277, ncol=349)
```

### Fill Empty Raster
Filling raster using rasterize function by assigning coordinates and the temperature value on 2015-01-01. For point values that may fall in to two grids, I've used the mean to average the two.
```{r fill_raster}
# new temp raster with latlon coords 
r_temp <- rasterize(temp_coords[,1:2], r, as.vector(t_temp), fun=mean)
```

### Extract Values to Points
Using the extract function to extract the values from the temperature raster to the grid locations of the PM grid. 
```{r extract}
# extract values to pm points
temp_k_regrid <- extract(r_temp, pm_coords)
```

### Plot Check
Plotting regridded temperature values on the PM grid.
```{r regrid_plot}
# creating dataframe
pm_temp_test <- data.frame(cbind(pm_coords, temp_k_regrid)) %>% 
  rename(lon = V1, lat = V2, temp_k = temp_k_regrid)

# plot
ggplot(pm_temp_test, aes(x=lon, y=lat, color=temp_k)) +
  geom_point() +
  theme_minimal()
```

This general process will work and I think it will be relatively efficient.The next step is to itterate this and consider parallel processing for multiple days and years.






