---
title: "Colorado Respiratory Morbidity Association"
author: "Ryan Gan"
date: "9/24/2018"
output: html_document
---

## Introduction

This is an R Markdown script prepared for Sheryl that contain pilot analyses evaluating a binary classifier of smoke exposure and the likelihood for a cardiopulmonary hospitalization (inpatient admission) admitted via the emergency room during the wildfire season of 2015 (May to September). The study population is limited to Colorado. Note, it is not limited to Front Range communities, so there may be some concern with exposure misclassification in areas without a PM2.5 monitor. The specific study design is the time-stratified case-crossover design with a referent period of 1-month.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries required to analyze data.

```{r library, echo=F, message=F}
library(tidyverse)
library(survival)
library(splines)
```

Load time-stratified case-crossover morbidity data.

```{r list_load, echo=F, message=F}
load('../../data/health/2015-morbidity_casecross_list.RData')
```

Filter list to only Colorado outcomes and remove larger list to save space. Creating some lagged variables to evaluate distributed lag relationships. I'm only looking at same-day to lagged day of 2. The rational for this is that smoke from the large fires or transported smoke in Colorado were not elevated as long as Washington or Oregon. I think a couple days should be adequate for a pilot evaluation.

```{r list_setup, echo=F, message=F}
co_list <- ts_casecross_list %>% 
  map(~filter(., state == 'colorado') %>% 
       # add lagged smoke vars
       mutate(smoke5_lag1 = ifelse(gwr_smk_lag1 >= 0.5, 1, 0),
              smoke5_lag2 = ifelse(gwr_smk_lag2 >= 0.5, 1, 0),
              smoke5_lag3 = ifelse(gwr_smk_lag3 >= 0.5, 1, 0),
              smoke10_lag1 = ifelse(gwr_smk_lag1 >= 1, 1, 0),
              smoke10_lag2 = ifelse(gwr_smk_lag2 >= 1, 1, 0),
              smoke10_lag3 = ifelse(gwr_smk_lag3 >= 1, 1, 0),
              smoke15_lag1 = ifelse(gwr_smk_lag1 >= 1.5, 1, 0),
              smoke15_lag2 = ifelse(gwr_smk_lag2 >= 1.5, 1, 0),
              smoke15_lag3 = ifelse(gwr_smk_lag3 >= 1.5, 1, 0)))
```

Same-day association for respiratory outcomes looking at smoke > 10 ug/m^3. I'm evaluating a cutoff of > 10 ug/m^3 because I think this better-captures smoke than a cutoff of >5, but it's also not as restrictively high as >15 ug/m^3. I think there are few events at that high cutoff. I will throughly explore this in the Colorado paper.

```{r sameday_results, echo=F, message=F}
# extract names
outcomes <- c('All Respiratory', 'Asthma', 'COPD', 'Acute Bronchitis',
              'Pneumonia', 'All CVD', 'Arrhythmia', 'Cerebrovascular',
              'Heart Failure', 'Ischemic Heart Disease', 
              'Myocardial Infarction')
# outcome order
out_order <- c('All Respiratory', 'Asthma', 'COPD', 'Acute Bronchitis',
              'Pneumonia', 'All CVD', 'Arrhythmia', 'Cerebrovascular',
              'Heart Failure', 'Ischemic Heart Disease', 
              'Myocardial Infarction')

same_day <- co_list %>% 
  map_dfr(., function(df){
    result <- broom::tidy(clogit(outcome ~ smoke10 + strata(identifier),
                                 data = df)) %>% 
      select(term, estimate, conf.low, conf.high) %>% 
      mutate_at(vars(estimate:conf.high), exp) 
  }) %>% 
  bind_cols(data.frame(outcomes), .) %>% 
  mutate(outcome = forcats::fct_relevel(outcomes, out_order),
         group = as.factor(if_else(outcomes %in% out_order[1:5],
          "Respiratory", "Cardiovascular")))
```

## Same-day plot all outcomes.

```{r sameday_plot, echo=F, message=F}
# plot
plot <- ggplot(data=same_day, aes(x=outcome, y = estimate, colour = group)) +
  geom_point() +
  geom_errorbar(aes(ymin=conf.low, ymax=conf.high), width = 0.3) +
  scale_color_manual("Cardiopulmonary", values = c("#ff00cc", "darkblue")) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  ylab(expression(paste("Odds Ratio: Smoke PM"[2.5], " > 10 ug/m"^3))) +
  xlab("Outcome") +
  theme_bw()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust=0.95, vjust = 0.75))

print(plot)
```

A couple notes for Sheryl in regard to this analysis that is Colorado only compared to the pooled Colorado and Washington:

1. I think the impact of smoke in Colorado was not as extreme as Washington so I need to think more about appropriate cutoffs. I think > 5 might be too low and capture a lot of anthropogenic events and > 15 might be too high for a lot of Colorado communities and would reduce the number of events in the exposed gorup. I settled on > 10. 

2. I think there may be some difference in response based on lagged days which I'll get in to below.

## Same-day association respiratory only

Sheryl has requested a respiratory hosptialization only plot since the grant is primarily focused on respiratory outcomes. 

```{r resp_only, echo=F, message=F}
# plot
plot <- ggplot(data=filter(same_day, group == 'Respiratory'), 
               aes(x=outcome, y = estimate)) +
  geom_point(color='darkblue') +
  geom_errorbar(aes(ymin=conf.low, ymax=conf.high), width = 0.3, 
                color = 'darkblue') +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  ylab(expression(paste("Odds Ratio: Smoke PM"[2.5], " > 10 ug/m"^3))) +
  xlab("Outcome") +
  theme_bw()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust=0.95, vjust = 0.75))

print(plot)
```

## Distributed lag results

Same-day associations show null or no assoication with respiratory hospitalizations when smoke is >10 ug/m^3. Hypothesis here could be that people are modifying their behavior and trying to avoid smoke or heeding air quality warnings. This would be a reason to look at lagged associations.I've used 2 degrees of freedom splines since there aren't that may days to lag over.

```{r dl_function, echo=F, message=F}
distribute_that_lag <- function(lag_mod, strata) {
  # output pm basis estimates
  parms <- broom::tidy(lag_mod) %>% 
    filter(stringr::str_detect(term, strata)) %>% 
    select(estimate) %>% 
    as_vector()
  # output estimate names for cov matrix
  names <- stringr::str_subset(names(lag_mod$coefficients), strata)
  # define lagged basis spline if it doesn't exist
  exp_b <- ns(0:2, df = 2, intercept = T)
  # estimate associations
  est <- exp_b %*% parms
  # estimate standard error for each interval
  # time variable
  time <- ((rep(1:length(est))-1))
  # covariance matrix for knots 
  cov_mat <- as.matrix(vcov(lag_mod))[names, names]
  # estimate variance of spline
  var <- exp_b %*% cov_mat %*% t(exp_b)
  # estimate lag ----
  # estimate standard error for each lag day for smoke
  l_se <- sqrt(diag(var))
  # calculate lower and upper bound for smoke
  l_est_l95 <- est + (l_se*qnorm(1-0.975))
  l_est_u95 <- est + (l_se*qnorm(0.975))
  l_type <- "lag"
  # lag dataframe
  l_df <- data.frame(strata, l_type, time, 
                     exp(est), exp(l_est_l95), exp(l_est_u95), 
                     row.names = NULL) 
  # assign column names
  colnames(l_df) <- c("strata", "type", "time", 
                      "odds_ratio", "lower_95", "upper_95")
  # cumulative estimates
  c_est <- sapply(seq_along(est), function(x){
    sum(est[1:x])
  })
  # stderr cumulative effect smk
  c_se <- sapply(seq_along(c_est), function(y){
    sqrt(sum(var[1:y,1:y]))
  })
  # estimate 95% CI
  c_l95 <- c_est+(c_se*qnorm(1-0.975))
  c_u95 <- c_est+(c_se*qnorm(0.975))
  # type
  c_type <- "cumulative"
  # return dataframe
  c_df <- data.frame(strata, c_type, time, exp(c_est), 
                     exp(c_l95), exp(c_u95), row.names = NULL) 
  # assign column names
  colnames(c_df) <- c("strata", "type", "time", 
                      "odds_ratio", "lower_95", "upper_95")
  # bind lagged and cumulative 
  lag_est <- rbind(l_df, c_df) %>% 
    mutate(strata = as.character(strata),
           type = as.character(type))
  # return lagged estimate
  return(lag_est)
} # end lag estimate function
```


```{r dl_results, echo=F, message=F}
out_rep <- rep(outcomes, each = 6)

dl_results  <- lapply(co_list, function(x){
  # output dataframe from list
  data <- x %>% 
    mutate(date = as.Date(date),
           outcome = as.numeric(as.character(outcome))) %>%
    # remove missing lagged data
    filter(!is.na(smoke10_lag2))

  # create lagged matrix
  pm_mat <- as.matrix(select(data, smoke10, smoke10_lag1, smoke10_lag2))

  # define lagged basis spline
  exp_b <- splines::ns(0:(ncol(pm_mat)-1), df = 2, intercept = T)
  # pm basis
  pm_basis <- pm_mat %*% exp_b
  # run lagged model
  lag_mod <- clogit(outcome ~ pm_basis  + strata(identifier), data = data)

  # estimate lag estimate
  lag_est <- distribute_that_lag(lag_mod, strata = "pm") 
  return(lag_est)
  }) %>%  #end lappply
  # bind rows
  map_dfr(.,rbind) %>% 
  # bind in outcome names
  bind_cols(data.frame(out_rep), .)
```

## Cumulative effect over 0 to 2 lagged days of smoke exposure > 10 ug/m^3

Here is the cumulative effect of 3 days of smoke exposure > 10 ug/m^3 on cardiopulmonary respiratory hospitalizations. The relationship is similar to the same day association in that there aren't really any 'significant' respiratory associations, but there is an increased risk for asthma hospitalizations. There are also associations with some cardiovascular outcomes.

I'd say that all respiratory and pneumonia might have something strange going on.

```{r cumulative_results, echo=F, message=F}
c_results <- dl_results %>% 
  filter(type == 'cumulative' & time == 2) %>% 
  mutate(out_rep = forcats::fct_relevel(out_rep, out_order),
         group = as.factor(if_else(out_rep %in% out_order[1:5],
          "Respiratory", "Cardiovascular")))

# plot
plot <- ggplot(data=c_results, aes(x=out_rep, y = odds_ratio, 
                                   colour = group)) +
  geom_point() +
  geom_errorbar(aes(ymin=lower_95, ymax=upper_95), width = 0.3) +
  scale_color_manual(values = c("#ff00cc", "darkblue")) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  ylab(expression(paste("Odds Ratio: Smoke PM"[2.5], " > 10 ug/m"^3))) +
  xlab("Outcome") +
  theme_bw()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust=0.95, vjust = 0.75))

print(plot)
```

## Cumulative Lagged Assoication Respiratory-specific.

```{r echo = F, message = F}
# plot
plot <- ggplot(data = filter(dl_results, type == 'cumulative' & 
                               time == 2 & (out_rep %in% out_order[1:5])), 
               aes(x=out_rep, y = odds_ratio)) +
  geom_point(color='darkblue') +
  geom_errorbar(aes(ymin=lower_95, ymax=upper_95), width = 0.3, 
                color = 'darkblue') +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  ylab(expression(paste("Odds Ratio: Smoke PM"[2.5], " > 10 ug/m"^3))) +
  xlab("Outcome") +
  theme_bw()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust=0.95, vjust = 0.75))

print(plot)
```

Sheryl, a couple things about this quick analysis is that it only contains results from 2015 as I had these exposure data before I had the rest of the smooth krig values from Kate. I will include all years when I analyze Colorado. However, those time-stratified case-crossover dataframes were not readily available and I'll need to make those. 

I can include all years if that's of interst though. Let me know.