---
title: "Asthma Distributed Lag"
author: "Ryan Gan"
date: "2018-03-13"
editor_options:
  chunk_output_type: inline
output:
  html_document:
    df_print: paged

---

```{r setup, include=F}
library(tidyverse)
library(survival)
library(splines)
library(lubridate)

# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, 
                      warning=FALSE, message=FALSE, echo=F)
```


## Introduction

Testing out distributed lag code on a case-crossover dataframe. Working with the case-crossover dataframe before scaling to other outcomes.

Reading in PM~2.5~ data.

```{r pm_data}
pm <- read_csv("../../data/smoke/1015-county_popwt_pm.csv") %>% 
  rename(pm_diff = pm_smk) %>% 
  mutate(day = as.factor(weekdays(date)),
         weekend = ifelse(day %in% c("Saturday", "Sunday"), 1, 0),
         month = as.factor(lubridate::month(date)),
         year = as.factor(lubridate::year(date)),
         season = as.factor(case_when(month %in% c(12, 1, 2) ~ "winter",
                                      month %in% c(3:5) ~ "spring",
                                      month %in% c(6:8) ~ "summer",
                                      month %in% c(9:11)~ "fall")),
         smoke10_hms = if_else(pm_diff > 10 & month %in% c(3:11) & hms > 0.1,
                               1, 0),
         smoke15_hms = if_else(pm_diff > 15 & month %in% c(3:11) & hms > 0.1,
                               1, 0)) %>% 
  select(-state) 
```

Oregon data did not come with a county indicator, so we used ZIPCODEs provided to find the county. Loading key created by Jingyang and assigning 5 digit fips variable.

```{r oregon_county_key}
# oregon county key 
oregon_county_key <- read_csv(paste0("../../data/census/",
                                    "oregon_zip_county_key.csv"),
                              col_types = cols(.default = "c")) 
# create 5 digit fips key
or_fips <- paste0("410", c(paste0("0",seq(1,9,2)), seq(11, 71, 2)))
# county key sorted by name
county <- sort(unique(oregon_county_key$county_name))
county_fips <- data.frame(or_fips, county) %>% mutate_all(as.character)

# combine zip and county, remove county name
or_zip_county <- oregon_county_key %>% 
  # bind fips value in
  rename(county = county_name) %>% 
  left_join(county_fips, by = "county") %>% 
  select(1,3) %>% 
  mutate_all(as.character)
```

Read in asthma data and join in Oregon fips.

```{r read_asthma}
asthma <- read_csv("../../data/health/1015-asthma_casecross", 
                   col_types = cols(.default = "c")) %>% 
  left_join(or_zip_county, by = "zip") %>% 
  # fill in missing fips values
  mutate(fips = if_else(is.na(fips), or_fips, fips)) %>% 
  # remove or_fips variable
  select(-or_fips) %>% 
  # remove missing colorado counties (i think military base)
  filter(!is.na(fips)) %>% 
  # convert some values
  mutate(outcome = as.numeric(outcome),
         date = as.Date(date),
         age = as.numeric(age),
         stay_length = as.numeric(stay_length)) %>% 
  left_join(pm, by = c("fips", "date")) %>% 
  filter(!is.na(pm_krig))
```

### Unadjusted Associations

Running conditional logistic regression models to account for within person design. Each observation has referent periods assigned within the same month as the event date. 

#### State-Specific Binary Smoke 10

Regressing binary event yes or no on binary smoke where the difference between estimated PM~2.5~ and seasonal median is > 10 ug/m^3 and HMS cover for the county is > 10%. Running by state to see if there is any state-level differences. 

```{r smoke10_state_association}
states <- as.factor(sort(unique(asthma$state))) 
state_strata <- sort(unique(asthma$state))

results <- state_strata %>% 
  map(~filter(asthma, state == .)) %>% 
  map_dfr(~ broom::tidy(clogit(outcome ~ smoke10_hms + strata(id), 
                          data = .))) %>% 
  select(estimate, conf.low, conf.high) %>% 
  mutate_all(exp) %>% 
  round(.,3) %>% 
  cbind(state_strata, .)

knitr::kable(results, 
  caption = "Odds ratio and 95% CI between Smoke10 and asthma for each state")
```

Looks like there is an association in Washington but no association in Colorado or Oregon. We've seen an association in Oregon when PM was assigned at the ZIPCODE level using the blended models that I believe were at a finer resolution. The outcome for Oregon is also slightly different where we had to match the inpatient hospitalizations admitted through the ER in Colorado and Washington. Oregon has no direct variable to do this, so Jingyang found observations by the same person that were within a day or two admitted first to the ER, then to inpatient.  

#### State-Specific Binary Smoke 15

Running conditional logistic regression models for binary smoke where the difference between estimated PM~2.5~ and seasonal median is > 15 ug/m^3 and HMS cover for the county is > 10%. Running by state to see if there is any state-level differences. 

```{r smoke15_state_association}
results <- state_strata %>% 
  map(~filter(asthma, state == .)) %>% 
  map_dfr(~ broom::tidy(clogit(outcome ~ smoke15_hms + strata(id), 
                          data = .))) %>% 
  select(estimate, conf.low, conf.high) %>% 
  mutate_all(exp) %>% 
  round(.,3) %>% 
  cbind(state_strata, .)

knitr::kable(results, 
  caption = "Odds ratio and 95% CI between Smoke15 and asthma for each state")
```

Association even stronger for Washington, but inverse associations for Colorado and Oregon. Could be real where people generally tried to avoid high PM days with visible smoke.

#### State-Specific Continous PM~2.5~

Evaluating continous PM~2.5~. 

```{r pm_state_association}
results <- state_strata %>% 
  map(~filter(asthma, state == .)) %>% 
  map_dfr(~ broom::tidy(clogit(outcome ~ pm_krig + strata(id), 
                          data = .))) %>% 
  filter(term == "pm_krig") %>% 
  select(estimate, conf.low, conf.high) %>% 
  mutate_all(exp) %>% 
  round(.,3) %>% 
  cbind(state_strata, .)

knitr::kable(results, 
  caption = "Odds ratio and 95% CI between PM and asthma for each state")
```

No associatoin between increases in PM~2.5~ at the county level and asthma.

### Distributed Lag Association

Setting up lagged dataframe for Washington. Creating a lag function.

```{r lagfun}
# defining a lag function
funlag <- function(var, n=7){
  var <- enquo(var)
  indices <- seq_len(n)
  map( indices, ~quo(lag(!!var, !!.x)) ) %>% 
    set_names(sprintf("%s_lag%d", rlang::quo_text(var), indices))
}
```

Creating lagged dataframe.

```{r distributed_lag}
# continous pm lagged dataframe
pm_lag <- pm %>% 
  filter(str_sub(fips,start=1,end=2) %in% c("08","53", "41")) %>% 
  select(fips, date, pm_krig, smoke10_hms, smoke15_hms) %>% 
  mutate(pm = pm_krig/10) %>% 
  arrange(fips, date) %>% 
  group_by(fips) %>% 
  mutate(., !!!funlag(pm,6), !!!funlag(smoke10_hms,6),
         !!!funlag(smoke15_hms,6))
```

Subset to just Washington so my models run faster. Joining lagged PM values to Washington state. 

```{r washington}
asthma_w <- asthma %>% 
  filter(state == "washington") %>% 
  select(-c(pm_krig, smoke10_hms, smoke15_hms)) %>% 
  left_join(pm_lag, by = c("fips", "date"))
```

Before I test for interaction, I'm going to start by evaluating the lagged relationship between PM~2.5~ and asthma. Perhaps misclassification at the county level requires accounting for the variation over a couple days before anything is observed? Just a thought. 

```{r lagged_matrix}
# create lagged matrix
pm_mat <- as.matrix(select(asthma_w, pm, contains("pm_lag")))
smk10_mat <- as.matrix(select(asthma_w, contains("smoke10_hms")))
smk15_mat <- as.matrix(select(asthma_w, contains("smoke15_hms")))
```

Defining a lagged exposure basis using 3 degrees of freedom and multiply the exposure basis by spline.

```{r define exposure basis}
# lagged basis spline
exp_b <- ns(0:6, df = 3, intercept = T)
# lagged pm x basis
pm_basis <- pm_mat %*% exp_b
# smk10 basis
smk10_basis <- smk10_mat %*% exp_b
# smk15 basis
smk15_basis <- smk15_mat %*% exp_b
```

#### Washington: Cumulative Lagged effect of PM~2.5~

Running distributed lag of PM~2.5~ without indicators of smoke.
```{r pm_lag_mod}
# pm model
mod <- clogit(outcome ~ pm_basis + temp_f + strata(id), data = asthma_w)
```

PM smoke estimates for lagged days.

```{r pm_lag_est}
# output pm basis estimates
pm_parms <- broom::tidy(mod) %>% 
  filter(stringr::str_detect(term, "pm")) %>% 
  select(estimate) %>% 
  as_vector()
# output estimate names for cov matrix
pm_names <- stringr::str_subset(names(mod$coefficients), "pm")

# estimate associations
pm_est <- exp_b %*% pm_parms
# cumulative estimates
pm_c_est <- sapply(seq_along(pm_est), function(x){
              sum(pm_est[1:x])
            })
# estimate standard error for each interval
# time variable
time <- ((rep(1:length(pm_est))-1))
# covariance matrix for knots 
pm_cov_mat <- as.matrix(vcov(mod))[pm_names, pm_names]
# estimate variance of spline
pm_var <- exp_b %*% pm_cov_mat %*% t(exp_b)
# stderr cumulative effect smk
pm_c_se <- sapply(seq_along(pm_c_est), function(y){
              sqrt(sum(pm_var[1:y,1:y]))
           })
# estimate 95% CI
c_pm_l95 <- pm_c_est+(pm_c_se*qnorm(1-0.975))
c_pm_u95 <- pm_c_est+(pm_c_se*qnorm(0.975))

# return dataframe
c_estimate <- data.frame(time, exp(pm_c_est), exp(c_pm_l95), 
  exp(c_pm_u95), row.names = NULL)
# assign column names
colnames(c_estimate) <- c("time", "odds_ratio", "lower_95", "upper_95")

knitr::kable(c_estimate, 
             caption = "Association between lagged PM2.5 and asthma")
```

#### Washington Cumulative Lagged Effect

Plot of cumulative effect of a 10 ug m^-3^ increase in PM~2.5~ on risk for asthma event using distributed lag in Washington state.

```{r pm_plot}
plot <- ggplot(c_estimate, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  ylab(expression("Odds Ratio: 10 ug/m^3 increase PM"[2.5])) +
  xlab("Lagged Days") +
  ggtitle("Lagged PM2.5 and relative association for asthma") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

#### Washington: Cumulative Lagged Effect of binary smoke 15 exposure

Evaluating binary smoke cutoff > 10 ug m^-3^.

```{r smk_lag_mod}
# pm model
mod <- clogit(outcome ~ smk10_basis + temp_f + strata(id), data = asthma_w)
```

PM smoke estimates for lagged days.

```{r smk_lag_est}
# output pm basis estimates
parms <- broom::tidy(mod) %>% 
  filter(stringr::str_detect(term, "smk")) %>% 
  select(estimate) %>% 
  as_vector()
# output estimate names for cov matrix
names <- stringr::str_subset(names(mod$coefficients), "smk")

# estimate associations
est <- exp_b %*% parms

# cumulative estimates
c_est <- sapply(seq_along(est), function(x){
              sum(est[1:x])
            })

# estimate standard error for each interval
# time variable
time <- ((rep(1:length(est))-1))
# covariance matrix for knots 
cov_mat <- as.matrix(vcov(mod))[names, names]
# estimate variance of spline
var <- exp_b %*% cov_mat %*% t(exp_b)
# stderr cumulative effect smk
c_se <- sapply(seq_along(c_est), function(y){
              sqrt(sum(var[1:y,1:y]))
           })

# estimate 95% CI
c_l95 <- c_est+(c_se*qnorm(1-0.975))
c_u95 <- c_est+(c_se*qnorm(0.975))

# return dataframe
c_estimate <- data.frame(time, exp(c_est), exp(c_l95), 
  exp(c_u95), row.names = NULL)
# assign column names
colnames(c_estimate) <- c("time", "odds_ratio", "lower_95", "upper_95")

knitr::kable(c_estimate, 
             caption = "Association between lagged smoke and asthma")
```

Looks like the association is significant on the day after.

```{r smoke_dl_plot}
plot <- ggplot(c_estimate, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  ylab(expression("Smoke (Yes vs No)")) +
  xlab("Lagged Days") +
  ggtitle("Lagged smoke and relative association for asthma") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

#### Washington: Distributed Lag Cumulative Effect Interaction Model

Setting up the PM matrix for smoke and non-smoke days.
```{r pm_nosmk_mat}
# smoke days
pm_smk_b <- pm_basis * smk10_basis 
# no smoke days
pm_nosmk_b <- pm_basis * ifelse(smk10_basis == 1, 0, 1)
```

Model with interaction, including smoke basis in the model.

```{r int_mod}
mod <- clogit(outcome ~ pm_smk_b + pm_nosmk_b + smk10_basis + temp_f +
                strata(id), data = asthma_w)
```

Making a function to calculate lagged association for each basis.

```{r dl_fun}
distribute_that_lag <- function(model, strata){
  # output pm basis estimates
  parms <- broom::tidy(mod) %>% 
    filter(stringr::str_detect(term, strata)) %>% 
    select(estimate) %>% 
    as_vector()
  # output estimate names for cov matrix
  names <- stringr::str_subset(names(mod$coefficients), strata)
  # estimate associations
  est <- exp_b %*% parms
  # cumulative estimates
  c_est <- sapply(seq_along(est), function(x){
                sum(est[1:x])
              })
  # estimate standard error for each interval
  # time variable
  time <- ((rep(1:length(est))-1))
  # covariance matrix for knots 
  cov_mat <- as.matrix(vcov(mod))[names, names]
  # estimate variance of spline
  var <- exp_b %*% cov_mat %*% t(exp_b)
  
  # stderr cumulative effect smk
  c_se <- sapply(seq_along(c_est), function(y){
                sqrt(sum(var[1:y,1:y]))
             })
  # estimate 95% CI
  c_l95 <- c_est+(c_se*qnorm(1-0.975))
  c_u95 <- c_est+(c_se*qnorm(0.975))
  # return dataframe
  c_estimate <- data.frame(strata, time, exp(c_est), exp(c_l95), 
    exp(c_u95), row.names = NULL)
  # assign column names
  colnames(c_estimate) <- c("strata", "time", "odds_ratio", "lower_95",
                            "upper_95")
  return(c_estimate)
} # end function
```

Estimating terms for PM in the presence of smoke, PM without smoke, and smoke.

```{r interaction_results}
int_results <- as.character(c("pm_smk", "pm_nosmk", "smk10")) %>% 
  map_dfr(~distribute_that_lag(mod, .))
```

Plot of each term in the interaction model.

```{r plot_dl1}
plot <- ggplot(int_results, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  facet_wrap(~strata, scales = "free_y") +
  ylab(expression("Odds Ratio")) +
  xlab("Lagged Days") +
  ggtitle("Washington") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

I've added the binary smoke basis in as well. I suppose if I were to interpret this, the effect of a 10 ug m^-3^ increase on non-smoke days, is inversely associated as PM increases, and this effect trends towards null effect as the over subsequent lagged exposure. PM on smoke days increases over time, but this trend is not significant. Same with binary smoke 10.

### Washington 2012 Interaction Model Fire Season

I am going to subset to the 2012 wildfire season (April through October) for Washington state to compare results to the 2012 Wildfire paper in GeoHealth.

```{r wa_2012}
wa_2012 <- asthma %>% 
  filter(state == "washington") %>% 
  filter(date >= "2012-03-01" & date <= "2012-11-01") %>% 
  select(-c(pm_krig, smoke10_hms, smoke15_hms)) %>% 
  left_join(pm_lag, by = c("fips", "date")) %>% 
  filter(!is.na(pm_lag6))
```

Looking at the same-day association between PM~2.5~ and asthma. Same-day association looks similar in effect size to the GeoHealth paper, even though the magnitude is about 14% vs 7%. Perhaps there was something unique about the Wenatchee Fire Complex, or perhaps the way PM was measured (more monitors -> better daily estimates of PM)? Or maybe it could be the time-frame? Perhaps I should limit estimates to April through October to eliminate high PM in winter?

```{r same_day_mod}
exposure <- c("pm", "smoke10_hms", "smoke15_hms")

results <-  exposure %>% 
  map_dfr(., ~ broom::tidy(clogit(as.formula(paste0("outcome ~",., 
    " + strata(id)")), data = wa_2012))) %>% 
  select(estimate, conf.low, conf.high) %>% 
  mutate_all(exp) %>% 
  round(.,3) %>% 
  cbind(exposure, .)

knitr::kable(results, 
  caption = "Odds ratio and 95% CI for PM and binary smoke and asthma")
```

Evaluating lagged interaction.

```{r lagged_interaction}
# create lagged matrix
pm_mat <- as.matrix(select(wa_2012, pm, contains("pm_lag")))
smk10_mat <- as.matrix(select(wa_2012, contains("smoke10_hms")))
# lagged basis spline
exp_b <- ns(0:6, df = 3, intercept = T)
# lagged pm x basis
pm_basis <- pm_mat %*% exp_b
# smk10 basis
smk10_basis <- smk10_mat %*% exp_b
# smoke days
pm_smk_b <- pm_basis * smk10_basis 
# no smoke days
pm_nosmk_b <- pm_basis * ifelse(smk10_basis == 1, 0, 1)

# interaction model
mod <- clogit(outcome ~ pm_smk_b + pm_nosmk_b + smk10_basis + 
                strata(id), data = wa_2012)
# lagged estimates
int_results <- as.character(c("pm_smk", "pm_nosmk", "smk10")) %>% 
  map_dfr(~distribute_that_lag(mod, .))

plot <- ggplot(int_results, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  facet_wrap(~strata, scales = "free_y") +
  ylab(expression("Odds Ratio")) +
  xlab("Lagged Days") +
  ggtitle("Washington 2012 Interaction Model") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

When I run an interaction model here, a couple things stick out. First, I don't think there is a difference in the effect size of PM~2.5~ on non-smoke days vs smoke days. The shape may be slightly different, over time.

### Colorado and Washington Interaction Model

Running this interaction model for Colorado and Washington to compare with my distributed lag results from the time-series models.

```{r co_wash}
asthma_lag <- asthma %>% 
  filter(state %in% c("colorado", "washington")) %>% 
  select(-c(pm_krig, smoke10_hms, smoke15_hms)) %>% 
  left_join(pm_lag, by = c("fips", "date")) %>% 
  filter(!is.na(pm_lag6))

# create lagged matrix
pm_mat <- as.matrix(select(asthma_lag, pm, contains("pm_lag")))
smk10_mat <- as.matrix(select(asthma_lag, contains("smoke10_hms")))
smk15_mat <- as.matrix(select(asthma_lag, contains("smoke15_hms")))
# lagged basis spline
exp_b <- ns(0:6, df = 3, intercept = T)
# lagged pm x basis
pm_basis <- pm_mat %*% exp_b
# smk10 basis
smk10_basis <- smk10_mat %*% exp_b
# smoke days
pm_smk_b <- pm_basis * smk10_basis 
# no smoke days
pm_nosmk_b <- pm_basis * ifelse(smk10_basis == 1, 0, 1)
# model
mod <- clogit(outcome ~ pm_smk_b + pm_nosmk_b + smk10_basis + temp_f +
                strata(id), data = asthma_lag)
# extract lagged results
int_results <- as.character(c("pm_smk", "pm_nosmk", "smk10")) %>% 
  map_dfr(~distribute_that_lag(mod, .))
# plot results
plot <- ggplot(int_results, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  facet_wrap(~strata, scales = "free_y") +
  ylab(expression("Odds Ratio")) +
  xlab("Lagged Days") +
  ggtitle("Washington and Colorado Interaction Model") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

### Washington Colorado April to October Interaction Model

Limiting years to only include months April through October.

```{r co_wa_apr_oct}
asthma_lag <- asthma %>% 
  filter(state %in% c("colorado", "washington")) %>% 
  filter(month %in% 4:10) %>% 
  select(-c(pm_krig, smoke10_hms, smoke15_hms)) %>% 
  left_join(pm_lag, by = c("fips", "date")) %>% 
  filter(!is.na(pm_lag6))
# create lagged matrix
pm_mat <- as.matrix(select(asthma_lag, pm, contains("pm_lag")))
smk10_mat <- as.matrix(select(asthma_lag, contains("smoke10_hms")))
smk15_mat <- as.matrix(select(asthma_lag, contains("smoke15_hms")))
# lagged basis spline
exp_b <- ns(0:6, df = 3, intercept = T)
# lagged pm x basis
pm_basis <- pm_mat %*% exp_b
# smk10 basis
smk10_basis <- smk10_mat %*% exp_b
# smoke days
pm_smk_b <- pm_basis * smk10_basis 
# no smoke days
pm_nosmk_b <- pm_basis * ifelse(smk10_basis == 1, 0, 1)
# model
mod <- clogit(outcome ~ pm_smk_b + pm_nosmk_b + smk10_basis + temp_f +
                strata(id), data = asthma_lag)
# extract lagged results
int_results <- as.character(c("pm_smk", "pm_nosmk", "smk10")) %>% 
  map_dfr(~distribute_that_lag(mod, .))
# plot results
plot <- ggplot(int_results, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  facet_wrap(~strata, scales = "free_y") +
  ylab(expression("Odds Ratio")) +
  xlab("Lagged Days") +
  ggtitle("Washington and Colorado Interaction Model") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

No association with PM on non-smoke days, but perhaps an association with PM on smoke days 3 or 4 days after. I do notice differences by state. Showing that here.

#### Washington April to October INteraction Model

Showing association for only Washington state. 
```{r wa_apr_oct}
asthma_lag <- asthma %>% 
  filter(state %in% c("washington")) %>% 
  filter(month %in% 4:10) %>% 
  select(-c(pm_krig, smoke10_hms, smoke15_hms)) %>% 
  left_join(pm_lag, by = c("fips", "date")) %>% 
  filter(!is.na(pm_lag6))
# create lagged matrix
pm_mat <- as.matrix(select(asthma_lag, pm, contains("pm_lag")))
smk10_mat <- as.matrix(select(asthma_lag, contains("smoke10_hms")))
smk15_mat <- as.matrix(select(asthma_lag, contains("smoke15_hms")))
# lagged basis spline
exp_b <- ns(0:6, df = 3, intercept = T)
# lagged pm x basis
pm_basis <- pm_mat %*% exp_b
# smk10 basis
smk10_basis <- smk10_mat %*% exp_b
# smoke days
pm_smk_b <- pm_basis * smk10_basis 
# no smoke days
pm_nosmk_b <- pm_basis * ifelse(smk10_basis == 1, 0, 1)
# model
mod <- clogit(outcome ~ pm_smk_b + pm_nosmk_b + smk10_basis + temp_f +
                strata(id), data = asthma_lag)
# extract lagged results
int_results <- as.character(c("pm_smk", "pm_nosmk", "smk10")) %>% 
  map_dfr(~distribute_that_lag(mod, .))
# plot results
plot <- ggplot(int_results, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  facet_wrap(~strata, scales = "free_y") +
  ylab(expression("Odds Ratio")) +
  xlab("Lagged Days") +
  ggtitle("Washington Interaction Model") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```
 
Both PM on smoke and non-smoke days likely increase risk for asthma hospitalizations, however the slope may be more pronounced and extreme for PM due to smoke 3 or 4 days after.
 
#### Colorado April to October Interaction MOdel

Colorado distributed lag interaciton models.
```{r co_apr_oct}
asthma_lag <- asthma %>% 
  filter(state %in% c("colorado")) %>% 
  filter(month %in% 4:10) %>% 
  select(-c(pm_krig, smoke10_hms, smoke15_hms)) %>% 
  left_join(pm_lag, by = c("fips", "date")) %>% 
  filter(!is.na(pm_lag6))
# create lagged matrix
pm_mat <- as.matrix(select(asthma_lag, pm, contains("pm_lag")))
smk10_mat <- as.matrix(select(asthma_lag, contains("smoke10_hms")))
smk15_mat <- as.matrix(select(asthma_lag, contains("smoke15_hms")))
# lagged basis spline
exp_b <- ns(0:6, df = 3, intercept = T)
# lagged pm x basis
pm_basis <- pm_mat %*% exp_b
# smk10 basis
smk10_basis <- smk10_mat %*% exp_b
# smoke days
pm_smk_b <- pm_basis * smk10_basis 
# no smoke days
pm_nosmk_b <- pm_basis * ifelse(smk10_basis == 1, 0, 1)
# model
mod <- clogit(outcome ~ pm_smk_b + pm_nosmk_b + smk10_basis + temp_f +
                strata(id), data = asthma_lag)
# extract lagged results
int_results <- as.character(c("pm_smk", "pm_nosmk", "smk10")) %>% 
  map_dfr(~distribute_that_lag(mod, .))
# plot results
plot <- ggplot(int_results, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  facet_wrap(~strata, scales = "free_y") +
  ylab(expression("Odds Ratio")) +
  xlab("Lagged Days") +
  ggtitle("Colorado Interaction Model") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

You can see that Colorado has some very weird shapes that really don't make too much sense. 

I couldn't even fit models with Oregon in a timely manner. 

### Exploring Possible Non-Linearity in PM Response

I'm going to try one more thing. I'm interested in non-linear associations with PM, and the distributed lag non-linear models can accomplish this, but that seems complicated. So I'm going to create some dummy variables of categories of PM instead. I'll look just at Washington.

```{r pm_cat_lag}
# pm category
pm_cat <- pm %>% 
  filter(str_sub(fips,start=1,end=2) %in% c("53")) %>% 
  select(fips, date, pm_krig) %>% 
  mutate(pm_cat = factor(case_when(pm_krig <= 10 ~ "0-10", 
    pm_krig > 10 & pm_krig <= 20 ~ "10-20",
    pm_krig > 20 & pm_krig <= 50 ~ "20-50",
    pm_krig > 50 ~ ">50"), levels = c("0-10", "10-20", "20-50", ">50")),
    pm_cat0 = ifelse(pm_cat == "0-10", 1, 0),
    pm_cat1 = ifelse(pm_cat == "10-20", 1, 0),
    pm_cat2 = ifelse(pm_cat == "20-50", 1, 0),
    pm_cat3 = ifelse(pm_cat == ">50", 1, 0)) %>% 
    arrange(fips, date) %>% 
    group_by(fips) %>% 
    mutate(., !!!funlag(pm_cat0,6), !!!funlag(pm_cat1,6),
         !!!funlag(pm_cat2,6), !!!funlag(pm_cat3,6))
  

# asthma cat
asthma_cat <- asthma %>% 
  filter(state %in% c("washington", "colorado")) %>% 
  filter(month %in% 4:10) %>% 
  select(-c(pm_krig, smoke10_hms, smoke15_hms)) %>% 
  left_join(pm_cat, by = c("fips", "date")) 

# model
mod <- clogit(outcome ~ pm_cat + strata(id), asthma_cat)
# output just odds ratio
plot_df <- broom::tidy(mod) %>% 
  select(term, estimate, conf.low, conf.high) %>% 
  mutate_at(vars(estimate:conf.high), exp) %>% 
  mutate(term = forcats::as_factor(term, c("pm_cat10-20", "pm_cat_20-50",
                                             "pm_cat>50")))

# plot odds ratio for categories
ggplot(plot_df, aes(x=term, y=estimate)) +
  geom_point() +
  ylab("odds ratio")
```

Association looks more or less linear as PM increases. I'll try out the distributed lag approach though.

```{r}
# dummy cat matrix
pm_cat1_mat <- as.matrix(select(asthma_cat, contains("pm_cat1")))
pm_cat2_mat <- as.matrix(select(asthma_cat, contains("pm_cat2")))
pm_cat3_mat <- as.matrix(select(asthma_cat, contains("pm_cat3")))

# lagged basis spline
exp_b <- ns(0:6, df = 3, intercept = T)
# lagged dummy cat basis
p_cat1_b <- pm_cat1_mat %*% exp_b
p_cat2_b <- pm_cat2_mat %*% exp_b
p_cat3_b <- pm_cat3_mat %*% exp_b

# lagged clogit
mod <- clogit(outcome ~ p_cat1_b + p_cat2_b + p_cat3_b + temp_f +
              strata(id), asthma_cat)

cat_est <- c("p_cat1_b", "p_cat2_b", "p_cat3_b") %>% 
  map_dfr(~distribute_that_lag(mod, strata = .)) %>% 
  mutate(pm_strata = factor(
    case_when(strata == "p_cat1_b" ~ "10-20 ug/m^3",
              strata == "p_cat2_b" ~ "20-50 ug/m^3",
              strata == "p_cat3_b" ~ ">50 ug/m^3"), 
    levels = c("10-20 ug/m^3", "20-50 ug/m^3", ">50 ug/m^3")))

# plot results
plot <- ggplot(cat_est, aes(x=time, y=odds_ratio)) +
  geom_line(colour = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "blue", alpha = 0.3) + 
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  facet_wrap(~pm_strata, scales = "free_y") +
  ylab(expression("Odds Ratio: Strata vs 0-10 ug/m^3 strata")) +
  xlab("Lagged Days") +
  ggtitle("Risk for Asthma at varying PM2.5 Category Model") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor = element_blank()) +
  theme_minimal()

print(plot)
```

