---
title: "Wildfire Smoke and Mortality "
author: Ryan Gan
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Introduction

This notebook contains code and methods to evaluate the association between wildfire smoke and inpatient via emergency room morbidity. I will be using daily county-level counts of cardiopulmonary emergency room visits that resulted in a hospital inpatient admission as my primary outcome.

### Set up

Setting up knitr chunk options.

```{r kniter_opt}
knitr::opts_chunk$set(fig.height = 6, fig.width = 8, fig.align = "center")
```

Loading the tidyverse package, which contains most packages I will need for data cleaning, visualizations, and analysis. Loading stringr package for character string manipulation.

```{r packages}
library(tidyverse)
library(stringr)
```

### Data Import

Importing the Colorado and Washington state inpatient emergency department visits.

#### Note 2018-01-29: 
Consider adding all inpatient visits as well for all these states. Would this be a general question of hospital utilization?

Importing Colorado 2010 to 2015 time series and viewing first observations.

```{r colorado_import}
# define relative path to data folder
co_path <- paste0("../../data/health/2010-2015_morbidity_co_ts.csv")
# read in dataframe using read_csv function from tidyverse package
co_ts <- read_csv(co_path) 
# glimpse first observations
glimpse(co_ts)
```

Importing Washington 2011 to 2015 time series and viewing first observations.

```{r washington_import}
# define relative path
wa_path <- paste0("../../data/health/2011-2015_morbidity_wa_ts.csv")
# read in dataframe
wa_ts <- read_csv(wa_path) %>% 
  mutate(fips = as.character(fips))
# glimpse first obs
glimpse(wa_ts)
```

Spot for Oregon import.

Importing population county counts.
