---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Introduction

This notebook contains code and methods of time-stratified case-crossover analyses to check robustness of results for the county-level time-series analysis.

## Set up

Setting up knitr chunk options. Hiding messages, warnings, and code so that the document is easier to read for collaborators.

```{r kniter_opt}
knitr::opts_chunk$set(fig.height = 6, fig.width = 8, fig.align = "center",
                      progress = F, message = F, warning = F, echo=F,
                      results = "asis")
```

Loading the tidyverse package, which contains most packages I will need for data cleaning, visualizations, and analysis. Loading stringr package for character string manipulation.

```{r packages}
library(tidyverse) # all-purpose packages
library(stringr) # working with character strings
library(case.crossover) # loading my case crossover function
library(survival)
library(splines) # splines package
```

Setting up dark plot color theme.

```{r ggtheme}
ryan_theme <- theme(panel.background = element_rect(fill = "black", 
          colour = "black", size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, colour = "white"),
        panel.grid.minor = element_line(size = 0.1, color = "white", 
                                        linetype = "dotted"),
        plot.background = element_rect(fill = "#141e30", colour="#141e30",
                                       size = 0.5, linetype = "solid"),
        text = element_text(colour = "white"),
        axis.text = element_text(colour = "white"),
        strip.background = element_blank(),
        strip.text = element_text(colour="white"),
        legend.background = element_blank(),
        legend.key = element_blank())
```

# Emergency Room Hospitalizations

