---
title: "Western US County Population-Weighting PM"
author: "Ryan Gan"
date: "2018-01-17"
output: html_document
---

## Purpose

This markdown file contains descriptions and code for estimating daily county-level population-weighted PM ~2.5~ values from gridded estimates of PM ~2.5~. Gridded estimates were calculated using geographically-weighted ridge regression (GWR) and krigging.

### Setup

Loading tidyverse for general data wrangling and plotting; netcdf4 for accessing PM data stored in .nc file; sf package for spatial calculations.

```{r setup, include=F}
# libraries used
library(tidyverse) 
library(ncdf4) # package for working with .nc files
library(sf)

# general knitr options
knitr::opts_chunk$set(fig.width = 7, fig.height = 5, 
                      message = F, warning = F)
```

## Extracting PM values from netcdf files

Connecting to PM netcdf file and printing summary. NetCDF file is a 3-dimentional grid of latitude, longitude, and time. It also contains values for GWR, kriging, and background estimates of PM (estimated via GWR).

```{r connect_nc}
# open connection to netcdf 
smk_2015_nc <- nc_open("../data/smoke/2015-smoke_pm_wus.nc")
# print summary
smk_2015_nc
```

Extracting grid points (matrix) and setting as vectors. When using the as.vector function on matrix values from a netcdf, it will go row by row (left to right, then top to bottom). 

```{r assemble_grid}
# extract longitude as a vector
lon <- ncvar_get(smk_2015_nc, "glons")
dim(lon)
# extract latitude as a vector
lat <- ncvar_get(smk_2015_nc, "glats")
# extract date as a vector (format yyyymmdd)
date <- ncvar_get(smk_2015_nc, "dates")

# vector of lat and lon
lon_vec <- as.vector(lon)
lat_vec <- as.vector(lat)
# id sequence
id <- 1:20838

# bind together and plot to see where grids are
id_location <- as_data_frame(cbind(id, lon_vec, lat_vec))
```

Plotting the latitude and longitude coordinates to check centroids of grid.

```{r check_grid_ids}
# last 100
id_toprow <- id_location[c(20563:20573, 20701:20710),]
# plot of start of two top rows
ggplot(id_toprow, aes(x=lon_vec, y=lat_vec)) +
  geom_text(aes(label = id), size = 2.5) +
  xlim(-128, -125) +
  ylim(49.5, 50)+
  ggtitle("NetCDF cell locations: upper lefthand") +
  theme_minimal()

# first 100 values 
id_bottomrow <- id_location[c(1:10, 139:149),]
# plot of start of two bottom rows
ggplot(id_bottomrow, aes(x=lon_vec, y=lat_vec)) +
  geom_text(aes(label = id), size = 2.5) +
  xlim(-123, -121) +
  ylim(min(lat), 29.7) +
  ggtitle("NetCDF cell locations: lower lefthand") +
  theme_minimal()
```

I need to align the way as.vector() extracts values from matrices in the NetCDF and how the shapefile WRFGRID_IDs are labeled.

Working with smaller example matrices to make sure grids align.

```{r test_mat}
# example of how netcdf lat lon grid is arranged
test_mat_cell_id <- t(apply(matrix(12:1, nrow=4, ncol=3, byrow = T),1,rev))
test_mat_cell_id
# example of how wrfgrid shapefile is arranged
test_shape_mat <- apply(matrix(1:12, nrow=4, ncol=3), 2, rev)
test_shape_mat

# creation of test key
test_key <- as_data_frame(cbind(as.vector(test_mat_cell_id),
                                as.vector(test_shape_mat))) %>% 
  rename(id = V1, WRFGRID_ID = V2)

# check test key
head(test_key)
```

Creating WRFGRID_ID variable that will match how values are extracted from the NetCDF file (bottom left corner, going left to right, bottom to top).

```{r wrfgrid_id}
# 138 columns, 151 rows
# recreating a matrix that matches values
grid_id <- t(apply(matrix(20838:1, nrow=151, ncol=138, byrow = T),1,rev))
# extracting values as a vector that matches grid values in netcdf
cell_id <- as.vector(grid_id)

# WRFGRID_ID matrix based on cell location
wrf_id_mat <- apply(matrix(1:20838, nrow=151, ncol=138), 2, rev)
# vector of WRFGRID_ID
WRFGRID_ID <- as.vector(wrf_id_mat)

# join WRFGRID_ID values to lat lon coordinates
wrf_id_key <- as_data_frame(cbind(cell_id, WRFGRID_ID)) %>% 
  rename(id = cell_id) %>% 
  left_join(id_location, by = "id") %>% 
  rename(lat = lat_vec, lon = lon_vec) %>% 
  mutate(combine_id = paste(id, WRFGRID_ID, sep = ":"))
```

Plots are checking the top and bottom lefthand corner to make sure grid IDs match the WRFGRID_ID variable in the shapefile.

```{r check_wrfid_plot}
# extract ids from upper lefthand corner
id_toprow2 <- wrf_id_key %>% 
  filter(id %in% c(20563:20573,20701:20710))
# plot to check
ggplot(id_toprow2, aes(x=lon, y=lat)) +
  geom_text(aes(label = combine_id), size = 2.5) +
  xlim(-128, -125) +
  ylim(49.5, 50)+
  ggtitle("Cell locations: upper lefthand (NetCDF ID : WRF ID)") +
  theme_minimal()

# check bottom row
id_bottomrow2 <- wrf_id_key %>% 
  filter(id %in% c(1:10, 139:149))
# plot
ggplot(id_bottomrow2, aes(x=lon, y=lat)) +
  geom_text(aes(label = combine_id), size = 2.5) +
  xlim(-123, -121) +
  ylim(min(lat), 29.7) +
  ggtitle("Cell locations: bottom lefthand (NetCDF ID : WRF ID)") +
  theme_minimal()
```

## Checking PM values

I am going to check that the PM values are spatially projected correctly by joining GWR PM values with the shapefile before I population-weight. Going to check dates 2015-08-02 and 2015-08-21 for the GWR model.

```{r smoke_plots}
# extracting gwr blended estimates for dates
# kriging values in an array (x=151, y=138, z=122)
gwr_pm <- ncvar_get(smk_2015_nc, "gwrPM")

# find indices for dates I want to check
which(date==20150802) # 63
which(date==20150821) # 82

# extract dates from array
gwr_0802 <- as.vector(gwr_pm[,,63])
gwr_0821 <- as.vector(gwr_pm[,,82])

# bind with ids
# bind together and plot to see where grids are
gwr_check <- as_data_frame(cbind(id, lon_vec, lat_vec, 
                                 gwr_0802, gwr_0821)) %>% 
  # join WRFGRID ID in
  left_join(wrf_id_key, by = "id") %>% 
  select(-lon_vec, -lat_vec)

# load sf objects
# fips code for states I want to look at
western_state_fp <- c("04","06","08","16","53","41","32","49","56",
                      "35", "30")
# read county grid
wus_sf <- st_read("../data/shapefile/us_state", 
                  layer="cb_2016_us_state_20m") %>% 
  filter(STATEFP %in% western_state_fp)

# extract crs
nad83 <- st_crs(wus_sf)

# read grid sf
grid_sf <- st_read("../data/shapefile/smoke_grid", layer="wus_grid2015-ko")
# set projection
st_crs(grid_sf) <- nad83

# bind in with sf file 
gwr_check_sf <- grid_sf %>% 
  left_join(gwr_check, by = "WRFGRID_ID") %>% 
  select(id, WRFGRID_ID, gwr_0802, gwr_0821) %>% 
  # transforming variables
  mutate(gwr_0802 = ifelse(gwr_0802 < 0, 0, gwr_0802),
         gwr_0821 = ifelse(gwr_0821 < 0, 0, gwr_0821))

# plot of PM over western US on 2015-08-21
plot <- ggplot(wus_sf) +
  geom_sf(color = "red", alpha = 0) +
  geom_sf(data=gwr_check_sf, aes(fill = gwr_0821), 
          color = "transparent", alpha = 0.8) +
    scale_fill_gradient2(expression("PM2.5 µg/m"^3),
      low = "white", mid= "#00dbde", high="#6a3093", midpoint = 60) +
  ggtitle("GWR PM2.5 on 2015-08-21") +
  theme_minimal()

plot
```

The projection of PM~2.5~ looks good and matches satellite and WRF-chem estimates the Bonne provided that show the Eastern portion of Washington, Idaho, and Montana were impacted by smoke on 2015-08-21.

## Setting up PM matrices

I need to set up PM matrices where each row represents a WRFGRID and each column represents the day of estimation. I have alread extracted the GWR array from the NetCDF file. I need to extract krig and background array from NetCDF. I will also need to make estimates of smoke matrices, which are the cell values of krig and GWR subtracting off median values of each cell over the 122 days when no smoke was detected in the atmospheric column (assessed via HMS satellites). 

I will use apply to extract the 3rd element of the array (date-specific values).

```{r pm_matrices}
# extract krig_pm and background_pm from nc file ----
krig_pm <- ncvar_get(smk_2015_nc, "kPM")
background_pm <- ncvar_get(smk_2015_nc, "gwr_background")

# create cell x date matrices ----
# apply function to extract pm values for each cell for each date as a vector
krig_pm_mat <- apply(krig_pm, 3, as.vector)
gwr_pm_mat <- apply(gwr_pm, 3, as.vector)
# background has only one value
background_pm_vec <- as.vector(background_pm) 

# check that matrices were created correctly
# dimensions look correct (20838 cells by 122 days)
dim(gwr_pm_mat)
# checking gwr matrix element column 82 to see if it matches values of gwr0821
identical(gwr_pm_mat[,82], gwr_0821)

# set values <0 to 0
krig_pm_mat <- ifelse(krig_pm_mat < 0, 0, krig_pm_mat)
gwr_pm_mat <- ifelse(gwr_pm_mat <0, 0, gwr_pm_mat)

# create smoke estimates -----
krig_pm_smk_mat <- krig_pm_mat - background_pm_vec
gwr_pm_smk_mat <- gwr_pm_mat - background_pm_vec
# set values lower than 0 to 0
krig_pm_smk_mat <- ifelse(krig_pm_smk_mat < 0, 0, krig_pm_smk_mat)
gwr_pm_smk_mat <- ifelse(gwr_pm_smk_mat < 0, 0, gwr_pm_smk_mat)
```

Now that I have the individual matrices set up, I will perform the same operations and functions on each of the krig and GWR matrices. I'm going to bind all the matrices in a list and run them through lapply functions. I need to appply WRFGRID IDs to the row names and dates to the column names and then sort by row id to make sure the values are in the correct order for when I population-weight using linear algebra.

```{r assign_rowcolnames}
# assign WRFGRID IDs as rownames
# extract vector of arranged WRFGRID IDs that match cell IDs for rownames
WRFGRID <- arrange(wrf_id_key, id)$WRFGRID_ID
# extract dates for column names
date_vec <- paste0("date", date)

# create list of pm values
pm_list <- list(krig_pm_mat, krig_pm_smk_mat, gwr_pm_mat, gwr_pm_smk_mat)
# assign names to pm_list
names(pm_list) <- c("krig", "krig_smk", "gwr", "gwr_smk")

# lapply row and column names and sort by WRFGRID Cell
pm_list <- lapply(pm_list, function(x){
  colnames(x) <- date_vec
  rownames(x) <- WRFGRID
  x <- x[order(as.numeric(row.names(x))),]
})
```

I'm going to check that the cell values are in the correct order. To do that, I'll check the vector of GWR estimates on 2015-08-21 from the map I created above with the vector of GWR on 2015-08-21 in the list. If these two vectors match, the sorting of the matrices in the list worked.

```{r pm_check}
# sorted id
# extract gwr values as a vector on 08-21 from sf file
gwr0821_check <- arrange(gwr_check_sf, WRFGRID_ID)$gwr_0821
# extract gwr values from list on 08-21
gwr0821_check2 <- pm_list[[3]][,82]
# remove names on vector
names(gwr0821_check2) <- NULL

# check first 6 values
head(gwr0821_check)
head(gwr0821_check2)
# check summary stats
summary(gwr0821_check)
summary(gwr0821_check2)
# check identical
identical(gwr0821_check, gwr0821_check2)
```

Checks show that the matrices are in the correct order and ready to be population-weighted.

## Population-Weighting

Now that the matrices are ready, I need to read in the proportion intersection file and population density file.

I'm going to read the population density file first and plot to make sure it looks like the grid correctly aligns with where most people live. I need to check the density unit; pretty sure it's persons per km^2

```{r popdensity}
# load population density per grid file
popgrid <- read_csv("../data/smoke/2015-smoke_popgrid_wus.csv") %>% 
  rename(WRFGRID_ID = GRID_ID)

# create population grid matrix/vector for use later
pop_vec <- popgrid$PopDensity

# joining with grid sf object
pop_check_sf <- grid_sf %>% 
  left_join(popgrid, by = "WRFGRID_ID") %>% 
  # filter out places without people to make plotting easier
  filter(PopDensity > 100)

# plot of PM over western US on 2015-08-21
plot <- ggplot(wus_sf) +
  geom_sf(color = "black", alpha = 0) +
  geom_sf(data=pop_check_sf, aes(fill = PopDensity), 
          color = "transparent") +
    scale_fill_gradient("Population Density km^2",
      low = "white", high="red") +
  ggtitle("Population Density 2015 of grids > 100 persons km^2") +
  theme_minimal()

plot
```

Looks like the population density map shows the major population centers in the Western US. I only plotted densities > 100 to elimate a lot of sparsely populated grids to plot faster.

Loading in proportion intersect values I calculated in the 2015-smoke_gridcounty_intersection script.

```{r load_propint}
# load proportion intersect file; default columns as numeric
propint <- read_csv("../data/smoke/2015-smoke_propint_wus.csv", 
                    col_types = cols(.default = "n"))

# save fips vector; leaving out wrfgrid id column
fips <- stringr::str_sub(colnames(propint[,-1]), start = 6)

# create proportion intersect matrix
pi_mat <- as.matrix(propint[,-1])
# dimensions of pi matrix
dim(pi_mat)
```

I saved a vector of FIPS (state and county) IDs that I will use later in labeling. Produced a 20838 (cell) by 414 (county) matrix.

### Note: enter formula here

### County population-weighted PM

Adding background PM vector to PM list so that I can population-weight all at once.

```{r updated_pm_list}
# background pm matrix
background_mat <- as.matrix(background_pm_vec)
# assign wrfgrid rownames and sort
rownames(background_mat) <- WRFGRID
# order by wrfgrid
background_mat <- as.matrix(background_mat[order(as.numeric(
  row.names(background_mat))),])
# colname of background median
colnames(background_mat) <- "median"

# recreating list using elements from other list to preserve variables/order
pm_list2 <- list(pm_list[[1]], pm_list[[2]], pm_list[[3]], 
                 pm_list[[4]], background_mat)
# adding new name
names(pm_list2) <- c("krig", "krig_smk", "gwr", "gwr_smk", "background")
```

Creating a list of daily county population-weighted PM estimates and background PM estimates.

```{r popwt_pm}
# estimating county population density and inverse for use in formula
# estimate of population density per county
popden_county <- t(pi_mat) %*% pop_vec
# calcuate the inverse of county population density
popden_county_inverse <- 1/popden_county

popwt_pm_list <- lapply(pm_list2, function(x){
  # multiply population vector by pm matrices
    pm_pop_mat <- pop_vec * x
  # multiply pm population matrix by county/grid intersection matrix 
  # produces a summed estimate of pm values in a county for each day
    county_pm_sum_mat <- t(pi_mat) %*% pm_pop_mat
  # multiple county summed PM values by inverse of county population density
  # produces county population weighted PM estimates for each day
    county_popwt_pm <- county_pm_sum_mat * as.vector(popden_county_inverse)
  
  # create dataframe and join in FIPS id
    pm_county_wt <- as_data_frame(cbind(fips, county_popwt_pm)) %>% 
  # convert characters to numeric
      mutate_at(vars(-fips), funs(as.numeric))
  # return dataframe
  return(pm_county_wt)
})
```

Plot of 2015-08-21 county-population weighted GWR estimates of PM.

```{r popwt_plot}
# read county sf and limit to western US
county_sf <- st_read("../data/shapefile/us_county", layer="us_county") %>% 
  mutate(fips = paste0(STATEFP, COUNTYFP)) %>% 
  filter(STATEFP %in% western_state_fp) %>% 
  # join gwr population-weighted values to sf object
  left_join(as_data_frame(popwt_pm_list$gwr), by = "fips")

# plot of county PM over western US on 2015-08-21
plot <- ggplot(wus_sf) +
  geom_sf(color = "black", alpha = 0) +
  geom_sf(data = county_sf, aes(fill = date20150821), 
          color = "transparent", alpha = 0.8) +
    scale_fill_gradient2(expression("PM2.5 µg/m"^3),
      low = "white", mid= "#00dbde", high="#6a3093", midpoint = 50) +
  ggtitle("County GWR PM2.5 on 2015-08-21") +
  theme_minimal()

plot
```

Plot of county population-weighted PM~2.5~ corresponds with the gridded plot.

### Writing PM2.5 files

Writing county population-weighted files.

```{r write_popwt, eval=F}
# define write path
write_path <- "../data/smoke/"

# writing lapply to write all files in the population wt list 
lapply(1:length(popwt_pm_list), function(x){
  # write_name
  write_name <- paste0("2015-smoke_county_popwt_",
                       names(popwt_pm_list[x]),".csv")
  # save csv file
  write_csv(popwt_pm_list[[x]], paste0(write_path,write_name))
})
# Note: I'm not sure why it printed out each tibble when it wrote the file
```