---
title: "Western US County Population-Weighting PM"
author: "Ryan Gan"
date: "2018-01-17"
output: html_document
---

## Purpose

This markdown file contains descriptions and code for estimating daily county-level population-weighted PM ~2.5~ values from gridded estimates of PM ~2.5~. Gridded estimates were calculated using geographically-weighted ridge regression (GWR) and krigging.

### Setup

Loading tidyverse for general data wrangling and plotting; netcdf4 for accessing PM data stored in .nc file; sf package for spatial calculations.

```{r setup, include=F}
# libraries used
library(tidyverse)
library(ncdf4)

# general knitr options
knitr::opts_chunk$set(fig.width = 7, fig.height = 5, 
                      message = F, warning = F)
```

## Extracting PM values from netcdf files

Connecting to PM netcdf file and printing summary. NetCDF file is a 3-dimentional grid of latitude, longitude, and time. It also contains values for GWR, kriging, and background estimates of PM (estimated via GWR).

```{r connect_nc}
# open connection to netcdf 
smk_2015_nc <- nc_open("./data/smoke/2015-smoke_pm_wus.nc")
# print summary
smk_2015_nc
```

Extracting grid points (matrix) and setting as vectors. When using the as.vector function on matrix values from a netcdf, it will go row by row (left to right, then top to bottom). Plotting the latitude and longitude coordinates to check centroids of grid.

```{r assemble_grid}
# extract longitude as a vector
lon <- as.vector(ncvar_get(smk_2015_nc, "glons"))
# extract latitude as a vector
lat <- as.vector(ncvar_get(smk_2015_nc, "glats"))
# extract date as a vector (format yyyymmdd)
date <- as.vector(ncvar_get(smk_2015_nc, "dates"))


```


