---
title: "DLM Example"
author: "Ander Wilson"
date: "6/2/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulate example data

```{r simdat}
n <- 100
lag_max <- 7

# simulate lagged PM (iid normal)
pm <- data.frame(l0 = rnorm(n))
for(l in 1:lag_max){
  pm[,paste0("l",l)] <- c(pm[-c(1),paste0("l",l-1)],rnorm(1))
}
pm <- as.matrix(pm)

# true DL function
DLfunc <- exp(-(0:lag_max)/2)

# simulate outcomes
y <- drop(pm%*%DLfunc) + rnorm(n)
```


## Setup DLM

Here we setup the spline basis for the DLM.  I use natrual splines as the default but B-splines can be good and many other approaches work. Usually the degrees of freedom are small, maybe 3 or 4.
```{r setupdlmbasis}
library(splines)

B <- ns(0:lag_max, df =4, intercept=TRUE)

# the nrow B should equal ncol(pm)
dim(B)
dim(pm)
```


## Estiamte DLM with a linear model

First premultiple the lagged PM matrix by the basis
```{r premultiply}
pmB <- pm%*%B
```

Now estimate using the premultiplied exposures as the predictors. You can use a GLM or mixed model.
```{r estimatedlm}
fit <- lm(y~pmB)
```

The estimated regression coefficients are meaningless. Do not interpret them.
```{r coefs}
coef(fit)
```


## Find the estimates of the DL function

The point estimate is as follows. Note that it is the same length as the number of lags.
```{r pointest}
dlparms <- grep("pmB",names(coef(fit)))
DLestimate<- data.frame(estimate=B%*%coef(fit)[dlparms])
```

The variance matrix for the DL evaluated at each lagged time is as follows.
```{r var}
DLvar <- B%*%vcov(fit)[dlparms,dlparms]%*%t(B)
```

The standard errors at each point are as follows.
```{r sd}
DLestimate$SE <- sqrt(diag(DLvar))
```


The pointwise confidence interval assuming a  is
```{r ci}
DLestimate$lower <- DLestimate$estimate + DLestimate$SE * qt(1-0.975, df=fit$df.residual)
DLestimate$upper <- DLestimate$estimate + DLestimate$SE * qt(0.975, df=fit$df.residual)
```

See estimates.
```{r DLestimates}
DLestimate
```


## Plot the DL function


```{r plt}
library(ggplot2)
DLestimate$lag <- 0:lag_max
DLestimate$truth <- DLfunc
p <- ggplot(DLestimate, aes(x=lag, y=estimate, ymin=lower, ymax=upper))
p <- p + geom_ribbon(alpha=.5) + geom_line(size=2)
p <- p + geom_line(aes(x=lag, y=truth), linetype=2, size=2)
p <- p + theme_bw()
p
```


## Cumulative effect
The cumulative effect is the sum of the effect over all days.

```{r cumulative}
cumulative <- sum(DLestimate$estimate)
cumulative
cumulative_se <- sqrt(sum(DLvar))
cumulative_se
cumulative_CI <- cumulative + cumulative_se * qt(c(1-0.975,0.975), df=fit$df.residual)
cumulative_CI
```