---
title: "WRF-Grid Shapefile"
author: "Ryan Gan"
date: "2018-01-10"
output: html_document
---

## Purpose

The purpose of this script is to create a shapefile of the WRF-grid from a netcdf file to use in population-weighting calculations.

```{r setup}
library(tidyverse)
library(ncdf4)
library(raster)
library(sf)

# knitr options
knitr::opts_chunk$set(fig.width=7, fig.height=5,
                      warning=FALSE, message=FALSE)
```

## NetCDF file

Open connection to netcdf file, GWRdata4health_West2015, provided by Kate and Bonne using the netcdf4 package.

```{r nc_connection}
# define path to .nc file
nc_path <- paste0("../data/smoke/GWRdata4health_West2015.nc")
# open connection to .nc file
gwr_nc <- nc_open(nc_path)
# print summary of gwr nc file
gwr_nc
```

## Plot of WRF grid centroids

Extract latitude and longitude values and plot the center point of each cell.

```{r lat_lon_summary}
# extract lat values
lat <- ncvar_get(nc=gwr_nc, varid="glats")
# extract lon values
lon <- ncvar_get(nc=gwr_nc, varid="glons")

# turn lat/lon matrices in to flat file
latlon_locations <- as_data_frame(cbind(as.vector(lon), as.vector(lat))) %>% 
  rename(lon = V1, lat = V2)

summary(latlon_locations)

# plot centroids
ggplot(data=latlon_locations, aes(x=lon, y=lat)) +
  geom_point(size=0.5) +
  theme_minimal()
```

The plot shows that the points are not evenly spaced in a rectangular grid. This explains the problem we had when trying and read it in using the raster package. I think assigning a projection like WGS84 or NAD83 will fix this.

## Create a WRF raster (grid) based on irregular points

Converting irregular latitude and longitude points to a raster.

```{r create_raster}
# point locations
latlon <- cbind(as.vector(lon), as.vector(lat))
# create an extent object
e <- extent(latlon)
e

# find the dims of lat/lon matrix
dim(lon)
dim(lat)
r <- raster(e, nrow=138, ncol = 151)

# rasterize
wrfgrid <- rasterize(latlon, r)
# summary of wrfgrid raster
wrfgrid
```

## Create WRF polygon/shapefile

Creating a polygon object from the irregular raster.

```{r create_poly}
# creating polygon
wrf_poly <- rasterToPolygons(wrfgrid)
plot(wrf_poly)
```

## Convert polygon to simple features

Converting WRF polygon to simplefeatures object and plotting over US county polygon. I like working with simple features better.

```{r creat_sf}
# read county shapefile
us_state <- read_sf("../data/shapefile/us_state")
# extract proj4string from us county
nad83 <- st_crs(us_state)
# summary of us_state
# filter out states and territories outside of hte US
us_state <- us_state %>% 
  filter(!(NAME %in% c("Alaska", "Hawaii", "Puerto Rico")))
         
# converting polygon to sf object
wrf_sf <- st_as_sf(wrf_poly)
# assign wgs84 projection
st_crs(wrf_sf) <- nad83

head(wrf_sf)

# plot sf object overlayed on county
ggplot(us_state) + 
  geom_sf(color = "red") +
  geom_sf(data=wrf_sf, aes(), alpha = 0.0) +
  theme_minimal()
```

Polygon of WRFGRID is also irregular. Plotted it over a shapefile of counties in the US. I'm not sure if it's right, but there are some holes in the northern part of the grid where no polygons for the WRFGRID exists. 

The overall projection matches Bonne's and Kate's, but I'm going to check for potential gaps at the top.

### Plot of Montana subset

Subsetting grid to Montana.

```{r montana_check}
# subset montana state
montana_sf <- us_state %>% 
  filter(NAME == "Montana")

# subset grid to montana
montana_grid <- wrf_sf[montana_sf,]

# plot sf object overlayed on county
ggplot(montana_sf) + 
  geom_sf(alpha = 0.5, color = "red") +
  geom_sf(data=montana_grid, aes(), alpha = 0.0) +
  theme_minimal()
```

There does appear to be some gaps in the shapefile in Montana, but this might not be a huge issue since we have no health data from Montana. I will check Washington since we do have 2015 health data from there.

## Plot of Washington subset

```{r washington_check}
# subset montana state
washington_sf <- us_state %>% 
  filter(NAME == "Washington")

# subset grid to montana
washington_grid <- wrf_sf[washington_sf,]

# plot sf object overlayed on county
ggplot(washington_sf) + 
  geom_sf(alpha = 0.5, color = "red") +
  geom_sf(data=washington_grid, aes(), alpha = 0.0) +
  theme_minimal()
```

This may be okay. I do have some potential gaps in the shapefile though. Usually along the edge. I could figure this out and fix it, but that will take some time. I'll use Bonne's shapefile grid since it does not have gaps.