---
title: "Meta Time Series Analysis"
author: "Ryan_Gan"
date: "May 8, 2017"
output: html_document
---

## Background

This is a pooled analysis of state and year specific time-series evaluating the association between wildfire smoke exposure and cardipulmonary outcomes in the Western United States. 

### Notable differences between Washington and Oregon:

*Health Data:* Washington CHARS data captures persons who visited a hosptial in the state. Oregon data is insurance based and captures insured persons billed for a medical visit in the state of Oregon. 

*Smoke Data:* Washington smoke estimates were in general more accurate when compared with surface monitor estimates comapred to Oregon.

```{r libaries, message = F, warning = F}
library(tidyverse)
```

Read in meta time series data.

```{r read data, message = F, warning = F}
meta_ts <- read_csv("./data/meta_timeseries.csv")
summary(meta_ts)
```

## Descriptives

First step is to determine if there are any notable differences between the states. Going to aggregate counties by state (averaging health outcomes and smoke) and split states to Washington and Oregon.

```{r data wrangle, message = F, warning = F}
# washington
wa_ts <- meta_ts %>% filter(state == "WA")
# oregon
or_ts <- meta_ts %>% filter(state == "OR")

# averaged state aggregate
state_avg_ts <- meta_ts %>% group_by(state, date) %>% 
  # remove columns I don't want to calculate
  select(-fips, -county) %>% 
  # add up outcomes in each county
  summarise()

glimpse(meta_ts)
.cols
```
