---
title: "Meta Time Series Analysis"
author: "Ryan_Gan"
date: "May 8, 2017"
output: html_document
---

## Background

This is a pooled analysis of state and year specific time-series evaluating the association between wildfire smoke exposure and cardipulmonary outcomes in the Western United States. 

### Notable differences between Washington and Oregon:

*Health Data:* Washington CHARS data captures persons who visited a hosptial in the state. Oregon data is insurance based and captures insured persons billed for a medical visit in the state of Oregon. 

*Smoke Data:* Washington smoke estimates were in general more accurate when compared with surface monitor estimates comapred to Oregon.

Both state datasets were limited to only emergency room visits (billed for Oregon, observed for Washington).

```{r libaries, message = F, warning = F}
library(tidyverse)
library(lme4)
library(broom)
```

Read in meta time series data.

```{r read data, message = F, warning = F}
meta_ts <- read_csv("./data/meta_timeseries.csv") %>% 
  # creating 10 unit increase var of geo smoke and day of week
  mutate(geo_smk10 = geo_smk_pm/10,
         day = weekdays(date)) %>% 
  # creating lag variables for a crude distributed lag
  group_by(county) %>% 
  # create smoke and lag variables
  mutate(smoke0 = ifelse(geo_smk_pm > 0, 1, 0),
         smoke5 = ifelse(geo_smk_pm > 5, 1, 0),
         smoke10 = ifelse(geo_smk_pm > 10, 1, 0),
         smoke15 = ifelse(geo_smk_pm > 15, 1, 0), 
         # lag variables
         geo_smk_lag1 = lag(geo_smk_pm, 1, order_by = county),
         geo_smk_lag2 = lag(geo_smk_pm, 2, order_by = county),
         geo_smk_lag3 = lag(geo_smk_pm, 3, order_by = county),
         geo_smk_lag4 = lag(geo_smk_pm, 4, order_by = county),
         geo_smk_lag5 = lag(geo_smk_pm, 5, order_by = county),
         geo_smk_lag6 = lag(geo_smk_pm, 6, order_by = county),
         geo_smk_lag7 = lag(geo_smk_pm, 7, order_by = county),
         geo_smk_lag8 = lag(geo_smk_pm, 8, order_by = county),
         geo_smk_lag9 = lag(geo_smk_pm, 9, order_by = county),
         geo_smk_lag10 = lag(geo_smk_pm, 10, order_by = county)) %>% 
  ungroup()

# smoke wave days (at least 2 consecutive days in the 98th percentile of PM)
quantile(meta_ts$geo_wt_pm, probs = c(0.95,0.98,0.99, 0.999))
# following Coco's smoke wave day, the 98th percentile for PM2.5 concentrations
# would be >26.9

smoke_wave <- meta_ts %>% 
  select(state, county, date, geo_wt_pm) %>% 
  mutate(high_pm_day = ifelse(geo_wt_pm > 26.9, 1, 0)) %>% 
  group_by(state, county) %>% 
  # identify consecutive days
  mutate(smoke_wave = ifelse(lag(high_pm_day, 
    order_by = county)==1 & high_pm_day ==1, 1, 0)) %>% 
  select(-high_pm_day, -geo_wt_pm)

# merge in smokewave day
meta_ts <- meta_ts %>% 
  right_join(smoke_wave, by = c("state", "county", "date"))


```

### Incorporated the smoke-wave day approach by Coco Liu

Comparing smoke-wave approach to our method of defining smoke days

```{r smoke days, message=F, warning=F}
# only 150 smoke wave days
xtabs(~smoke_wave,  meta_ts)
# if we compare to our geo smoke > 10, we'd add 362 extra days
xtabs(~smoke_wave + smoke10, meta_ts)
# if we used smoke >15 we'd add 164 extra days 
xtabs(~smoke_wave + smoke15, meta_ts)

```

## Descriptive Plots

First step is to determine if there are any notable differences between the states. Going to aggregate counties by state (averaging health outcomes and smoke) and split states to Washington and Oregon.

```{r data wrangle, message = F, warning = F}

# washington
wa_ts <- meta_ts %>% filter(state == "WA")
# oregon
or_ts <- meta_ts %>% filter(state == "OR")

# count of state outcome timeseries
state_outcome_rate <- meta_ts %>% group_by(state, date) %>% 
  # remove columns I don't want to calculate
  select(-fips, -county) %>% 
  # add up outcomes in each county
  summarise_each(funs(sum), population, n_obs:broken_arm_n) %>% 
  mutate_each(funs((./population*100000)), n_obs:broken_arm_n) 

# average of environmental measures, joined with outcomes
state_smoke <- meta_ts %>% group_by(state, date) %>% 
  select(-fips, -county) %>% 
  summarise_each(funs(mean, median, min, max), wrf_pm:wrf_temp) %>% 
  full_join(state_outcome_count, by = c("state", "date"))

```

### Visualizing Exposure Time Series

Small multiples of smoke exposure series by smoke estimation method for each county within each state.

####Washington 2012 Smoke Time Series

```{r washington smoke, message = F, warning = F, results = "asis"}

# small multiples dataset
wa_smk <- wa_ts %>% 
  select(county, date, wrf_smk_pm, krig_smk_pm, geo_smk_pm) %>% 
  gather(key = smk_method, value = pm, -date, - county)
  
  
wa_smk_plot <- ggplot(wa_smk, aes(x = date, y = pm)) +
  geom_point(aes(colour = smk_method), alpha = 0.8, size = 0.8) +
  ggtitle("Washington: 2012 Fire Season Smoke PM2.5") +
  facet_wrap(~county) +
  ylab("Smoke PM2.5 ug/m^3") +
  xlab("Year: 2012") +
  theme_bw()

wa_smk_plot

```

####Oregon 2013 Smoke Time Series

```{r oregon smoke, message = F, warning = F, results = "asis"}

# small multiples dataset
or_smk <- or_ts %>% 
  select(county, date, wrf_smk_pm, krig_smk_pm, geo_smk_pm) %>% 
  gather(key = smk_method, value = pm, -date, -county)
  
  
or_smk_plot <- ggplot(or_smk, aes(x = date, y = pm)) +
  geom_point(aes(colour = smk_method), alpha = 0.8, size = 0.8) +
  facet_wrap(~county) +
  ggtitle("Oregon: 2013 Fire Season Smoke PM2.5") +
  ylab("Smoke PM2.5 ug/m^3") +
  xlab("Year: 2013") +
  theme_bw()

or_smk_plot

```

Seems like the southwest portion of Oregon was affected by smoke where the rest of the state did not have much smoke exposure. The smoke exposure timeframe also appears to have a smaller window of time on which it affected the area.

### Visualizing Health Outcome Time Series

Comparison between Oregon and Washington aggregated health outcomes over time.

```{r state outcome trends, message = F, warning = F, results = "asis"}
# I want CVD and respiratory rates by each state so I can make a 4 panel
state_cvd_resp <- state_outcome_rate %>% 
  select(state, date, cvd_n, resp_n) %>% 
  gather(key = outcome, value = n, -date, -state)
  

count_plot <- ggplot(state_cvd_resp, aes(x = date, y = n)) +
  geom_point() +
  facet_wrap(~state+outcome, scales = "free") +
  ylab("Outcome Rate Per 100k Person") +
  xlab("Date")
  theme_bw()


count_plot
```

CVD rates in Oregon for 2013 and Washington 2012 were similar, with no clear trend in the outcome rate over time. 

All respiratory outcomes were slightly different between Washington and Oregon. It looks like Oregon has almost twice the respiratory ER rate compared to Washington. This could mean some difference in respiratory rates between states, or more likely could have something to do with the different data sources. I'll need to think about this more. Justifies accounting for state though. Also, both states appear to have trends in respiratory ER visits where there is a decline in summer months, with an increase in the rate around the fall. Oregon has a pretty pronounced seasonal variation. 

It would be good to take a look at state-specific outcome time-series.

###Washington Health Outcomes Time Series

```{r wash health, warning=F, message=F, results='asis'}
glimpse(wa_ts)
# small multiples dataset
wa_health <- wa_ts %>% 
  select(county, date, population, n_obs:broken_arm_n) %>% 
  gather(key = outcome, value = events, -date, -county, -population) %>% 
  mutate(rate_per_100k = (events/population)*100000) %>% 
  filter(outcome == "resp_n" | outcome == "cvd_n")

#glimpse(wa_health)

# plot of rates
count_plot <- ggplot(wa_health, aes(x=date, y=events)) +
  geom_point(aes(colour = outcome), alpha = 0.8, size = 0.8) +
  facet_wrap(~county, scales = "free") +
  ggtitle("Washington: 2012 Fire Season Respiratory and CVD Rates") +
  ylab("Rates Per 100k") +
  xlab("Year:2012") +
  theme_bw()

count_plot
```


*Oregon*

```{r oregon health, message=F, warning=F, results='asis'}
# small multiples dataset
or_health <- or_ts %>% 
  select(county, date, population, n_obs:broken_arm_n) %>% 
  gather(key = outcome, value = events, -date, -county, -population) %>% 
  mutate(rate_per_100k = (events/population)*100000) %>% 
  filter(outcome == "resp_n" | outcome == "cvd_n")

# plot of rates
count_plot <- ggplot(or_health, aes(x=date, y=events)) +
  geom_point(aes(colour = outcome), alpha = 0.8, size = 0.8) +
  facet_wrap(~county, scales = "free") +
  ggtitle("Oregon: 2013 Fire Season Respiratory and CVD Rates") +
  ylab("Rates Per 100k") +
  xlab("Year:2013") +
  theme_bw()

count_plot

```



### Scatter Plot of Outcome Rates and Geo Smoke PM~2.5~

Visualizing rates of outcomes by 10ug/m^3 increasing Geoweighted smoke PM~2.5~.

```{r scatter plot, message=F, warning=F, results="asis"}
# calclating rates per 100k
rate_df <- meta_ts %>% mutate_at(
    which(colnames(meta_ts)=="n_obs"):which(colnames(meta_ts)=="broken_arm_n"),
    funs((./population)*100000)) %>% 
  mutate_at(c("wrf_smk_pm", "krig_smk_pm", "geo_smk_pm"),
    funs(./10)) %>% 
  select(state, county, date, geo_smk_pm, geo_wt_pm,
  which(colnames(meta_ts)=="n_obs"):
  which(colnames(meta_ts)=="broken_arm_n")) %>% 
  gather(key = outcome, value = rate_100k, -state, 
         -county, -date, -geo_smk_pm, -geo_wt_pm) %>% 
  filter(outcome != "n_obs")
  

# plot of pm2.5 every 10 units and asthma
ggplot(rate_df, aes(x = geo_smk_pm, y = log(rate_100k+0.1))) +
  geom_point(aes(colour = state), alpha = 0.2,  size = 1) + 
  scale_colour_manual(values = c("red", "blue")) +
  # truncated y axis
  #scale_y_continuous(limits = c(0, 1.5)) +
  # fit line for Washington
  geom_smooth(data = filter(rate_df, state == "WA"), 
              method=lm, formula = y~x, se = F, colour = "blue") +
  # fit line for Oregon
  geom_smooth(data = filter(rate_df, state == "OR"),
              method=lm, formula = y~x, se = F, colour = "red") +
  facet_wrap(~outcome, scale  = "free") +
  ylab("Log Outcome Rate per 100,000 Persons") +
  xlab("GWR Smoke PM2.5 per 10ug/m^3") +
  theme_bw()  

```

It looks like there is a linear relationship with smoke and asthma (expected), maybe with cerebrovascular disease, COPD, and respiratory outcomes.

### Standard Regression Models

Same-day estimates of association between smoke via GWR and outcomes.

```{r standard analysis, message=F, warning=F, results="asis"}
# create empty dataframe
rr_df <- data.frame(matrix(nrow = 12, ncol = 4))
colnames(rr_df) <- c("outcome", "rel_risk", "lower_95", "upper_95")

# vector of outcomes
outcome_list <- colnames(meta_ts[, which(colnames(meta_ts)=="resp_n"):
    which(colnames(meta_ts)=="broken_arm_n")])

# names 
outcome_names <- c('All Respiratory', 'Asthma', 'COPD', 'Pneumonia', 
  'Acute Bronchitis', 'Cardiovascular Disease', 'Arrhythmia', 
  'Cerebrovascular Disease', 'Heart Failure', 'Ischemic Heart Disease',
  'Myocardial Infarction', 'Broken Arm')

# populate outcome column
rr_df[,1] <- factor(outcome_names, levels=unique(outcome_names))

meow <- lapply(outcome_list, function(x){

 smk_est <- tidy(glm(as.formula(paste0(x, 
    "~ geo_smk10 + wrf_temp + state + county + 
    day + offset(log(population))")), 
    meta_ts, family="quasipoisson"))[2, ]
 
 rr <- c(round(exp(smk_est[1,2]),3),
         round(exp(smk_est[1,2]-1.96*smk_est[1,3]),3),
         round(exp(smk_est[1,2]+1.96*smk_est[1,3]),3))
 
}) # end apply

for(i in 1:nrow(rr_df)){
  rr_df[i,2:4] <- meow[[i]]
}

# add in a grouping variable
rr_df$group <- NA
rr_df[1:5, 5] <- "Respiratory"
rr_df[6:11, 5] <- "CVD"
rr_df[12, 5] <- "Broken Arm"

# tried apply function and it might have been easier to just use a loop
# as I needed a for loop to fill the list in to the dataframe
# code is more compact though
# I should check speed (or make a custom function to do this one day)

ggplot(rr_df, aes(x=outcome, y = rel_risk, colour = group)) +
  geom_point() +
  geom_errorbar(aes(ymin=lower_95, ymax=upper_95), width = 0.2) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  theme(
   panel.background = element_rect(fill = 'white', colour = 'black'),
   panel.grid.major = element_blank(),
   panel.grid.minor = element_blank(),
   axis.text.x = element_text(angle = 90, vjust = .75)) +
  ylab("Relative Risk") +
  xlab("Outcome") +
  ggtitle("Relative Risk for a 10ug/m^3 increase PM2.5 estimated via GWR Smoke")
```

Smoke-wave associations with outcomes.

```{r standard analysis smoke wave, message=F, warning=F, results="asis"}
# create empty dataframe
rr_df <- data.frame(matrix(nrow = 12, ncol = 4))
colnames(rr_df) <- c("outcome", "rel_risk", "lower_95", "upper_95")

# vector of outcomes
outcome_list <- colnames(meta_ts[, which(colnames(meta_ts)=="resp_n"):
    which(colnames(meta_ts)=="broken_arm_n")])

# names 
outcome_names <- c('All Respiratory', 'Asthma', 'COPD', 'Pneumonia', 
  'Acute Bronchitis', 'Cardiovascular Disease', 'Arrhythmia', 
  'Cerebrovascular Disease', 'Heart Failure', 'Ischemic Heart Disease',
  'Myocardial Infarction', 'Broken Arm')

# populate outcome column
rr_df[,1] <- factor(outcome_names, levels=unique(outcome_names))

meow <- lapply(outcome_list, function(x){

 smk_est <- tidy(glm(as.formula(paste0(x, 
    "~ smoke_wave + wrf_temp + state + county + 
    day + offset(log(population))")), 
    meta_ts, family="quasipoisson"))[2, ]
 
 rr <- c(round(exp(smk_est[1,2]),3),
         round(exp(smk_est[1,2]-1.96*smk_est[1,3]),3),
         round(exp(smk_est[1,2]+1.96*smk_est[1,3]),3))
 
}) # end apply

for(i in 1:nrow(rr_df)){
  rr_df[i,2:4] <- meow[[i]]
}

# add in a grouping variable
rr_df$group <- NA
rr_df[1:5, 5] <- "Respiratory"
rr_df[6:11, 5] <- "CVD"
rr_df[12, 5] <- "Broken Arm"

# tried apply function and it might have been easier to just use a loop
# as I needed a for loop to fill the list in to the dataframe
# code is more compact though
# I should check speed (or make a custom function to do this one day)

ggplot(rr_df, aes(x=outcome, y = rel_risk, colour = group)) +
  geom_point() +
  geom_errorbar(aes(ymin=lower_95, ymax=upper_95), width = 0.2) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  theme(
   panel.background = element_rect(fill = 'white', colour = 'black'),
   panel.grid.major = element_blank(),
   panel.grid.minor = element_blank(),
   axis.text.x = element_text(angle = 90, vjust = .75)) +
  ylab("Relative Risk") +
  xlab("Outcome") +
  ggtitle("Relative risk for exposure to a smoke-wave")
```

### Crude Mediation

```{r mediation}
smk_pm <- lmer(geo_wt_pm ~ smoke_wave + wrf_temp + (1|day) +
                     (1|county) + (1|state), meta_ts)
# smoke is associated with higher pm2.5
summary(smk_pm)

# smoke is associated with asthma
smk_asthma <- glmer(asthma_n ~ smoke_wave + wrf_temp + (1|day) +
                     (1|county) + (1|state) +
                     offset(log(population)), 
                     meta_ts, family="poisson")
summary(smk_asthma)

# add in geo wt pm
mediation <- glmer(asthma_n ~ smoke_wave + geo_wt_pm + 
                     wrf_temp + (1|day) +
                     (1|county) + (1|state) +
                     offset(log(population)), 
                     meta_ts, family="poisson")

summary(mediation)


```


### Interaction Model

```{r interaction}
smoke_pm_int <- glmer(asthma_n ~ smoke15 + geo_wt_pm + smoke10:geo_wt_pm + 
                     wrf_temp + (1|day) +
                     (1|county) + (1|state) +
                     offset(log(population)), 
                     meta_ts, family="poisson")

summary(smoke_pm_int)


```

### Mixed Model

```{r mixed mod analysis, message=F, warning=F, results="asis", eval = F}
mixed_mod <- glmer(asthma_n ~ geo_wt_pm + wrf_temp + (1|day) +
                     (1|county) + (1|state) +
                     offset(log(population)), 
                     meta_ts, family="poisson")

summary(mixed_mod)

```


### Distributed Lag With Covariates in Model

This is likely not ideal, but may be a start

```{r asthma dist lag analysis, message=F, warning=F, results="asis", eval = F}
# I need to concatnate the state and counties as there are some counties with
# the same name in both states
meta_ts2 <- meta_ts %>%
  mutate(state_county = paste(state, county, sep = "_"))

glimpse(meta_ts2$state_county)


dis_lag_mod <- glmer(arrythmia_n ~ smoke_wave +
  wrf_temp + (1|day) + (1|county) + (1|state) +  
  offset(log(population)), meta_ts, family="poisson")

summary(dis_lag_mod)

```